---
title: "Ankle draft risk model under winter clothing conditions"
authors:
  - name: Tobias Kramer
    affiliations:
      - ref: cbe
    corresponding: true
    email: t.kramer@berkeley.edu
    orcid: 0000-0001-6250-6718
  - name: Junmeng Lyu
    affiliations:
      - ref: cbe
      - ref: sjtu
    orcid: 0000-0003-0043-4127
  - name: Stefano Schiavon
    affiliations:
      - ref: cbe
      - ref: arch
      - ref: eng
    orcid: 0000-0003-1285-5682 

affiliations:
  - id: cbe
    name: Center for the Built Environment, University of California, Berkeley, USA
  - id: arch
    name: Department of Architecture, College of Environmental Design, University of California, Berkeley, USA
  - id: eng
    name: Department of Civil and Environmental Engineering, College of Engineering, University of California, Berkeley, USA
  - id: sjtu
    name: Department of Architecture, School of Design, Shanghai Jiao Tong University, Shanghai, China

filters:
  - authors-block

date: "`r Sys.Date()`"
format:
  docx:
    link-citations: true
    reference-doc: template.docx
    toc: false
    number-sections: true
  html:
    toc: true
    number-sections: true
    theme: cosmo
    comments:
      hypothesis: true
bibliography: refs.bib
csl: building-and-environment.csl
execute:
  echo: false
  warning: false
  error: false
---

```{r load-main-analysis}
#| message: false
#| warning: false
#| paged-print: false
source(here::here("src", "R", "x_analysis.R"))
```

# Abstract {.unnumbered}

The ankle draft model in ASHRAE Standard 55 was derived from studies
where participants mainly wore summer clothing with exposed ankles
(\~0.6 clo), potentially leading to overly conservative perimeter
heating or facade insulation requirements in winter conditions where
occupants typically wear long trousers, socks, and closed-toe shoes
(\~0.75 clo). We evaluated thermal comfort responses to ankle-level
draft under these conditions in controlled laboratory experiments with
51 participants. In a within-subjects repeated-measures design,
participants were exposed to nine conditions varying in ankle-level air
temperature (17–20°C) and air speed (0.1–0.7 m/s), simulating window
downdraft.

The existing ankle draft model systematically overpredicted
dissatisfaction across all conditions. While prior studies with exposed
ankles reported dissatisfaction rates of 23–57%, we observed only 6–20%,
with the existing model overestimating by an average of 2.3 times and by
as much as 30% at high air speeds. We developed separate generalized
linear mixed-effects models for covered- and uncovered-ankle conditions
from this and prior studies. The covered-ankle model showed lower
baseline dissatisfaction risk and reduced sensitivity to air speed. Only
whole-body thermal sensation and ankle-level air speed were significant
predictors; ankle air temperature and room temperature were
non-significant within tested ranges. These results are subject to a
20-minute exposure duration and the exclusion of radiant asymmetry from
cold window surfaces.

These findings support differentiating ASHRAE Standard 55 ankle draft
criteria by clothing level. Adopting a covered-ankle criterion could
enable meaningful reductions in perimeter heating and facade insulation
without compromising occupant comfort, with benefits for energy use and
carbon emissions.

# Keywords {.unnumbered}

thermal comfort; ankle draft; local thermal discomfort; winter clothing;
ASHRAE Standard 55

# Highlights {.unnumbered}

-   Existing ankle draft model overpredicts dissatisfaction 2.3× under
    winter clothing
-   Covered-ankle dissatisfaction ranged 6–20% vs. 23–57% for exposed
    ankles
-   New covered-ankle model shows lower baseline risk and reduced air
    speed sensitivity
-   Only whole-body thermal sensation and air speed are significant
    draft predictors
-   Separate ASHRAE 55 ankle draft criteria by clothing level could
    reduce energy use

# Graphical abstract {.unnumbered}

![Graphical
abstract.](figs/graphical-abstract.png){#fig-graphical-abstract
width="100%"}

# Introduction

Cold downdraft near windows is a common source of local thermal
discomfort in buildings during winter [@heiselbergDraughtRiskCold1994;
@fangerAirTurbulenceSensation1988]. When outdoor temperatures drop, the
interior surface of windows, particularly large or poorly insulated
glazing, becomes significantly cooler than the surrounding air. This
temperature differential causes nearby air to cool, increase in density,
and descend, creating a localized airflow pattern that can cause
discomfort at the lower extremities of seated occupants
[@schellenDowndraughtAssessmentDesign2012;
@larssonExperimentalInvestigationDowndraught2002]. Building designers
typically address this issue by increasing the insulation level of
windows, so that the interior surface temperature is higher, by adding
mullions to break the downward flow or by installing perimeter heating
systems, such as radiators or in-floor heating, to counteract the
downdraft effect and to prevent thermal discomfort in the ankle region.

The current approach to evaluating ankle draft risk relies on a
predictive model presented in @liuPredictedPercentageDissatisfied2017,
which estimates the predicted percentage dissatisfied (PPD) with ankle
draft as a function of air speed and temperature at ankle level, as well
as whole-body thermal sensation, all factors previously identified as
key modulators of draft sensitivity
[@toftumDraughtSensitivityInfluenced1996;
@oguroEvaluationEffectAir2001]. This model was derived from laboratory
experiments conducted under cooling conditions typical of underfloor air
distribution (UFAD) and displacement ventilation systems and is
incorporated into ASHRAE Standard 55 [@ashraeANSIASHRAEStandard2023].
Importantly, the 110 participants in these foundational studies,
including an earlier investigation by
@schiavonSensationDraftUncovered2016, wore summer clothing with exposed
ankles, resulting in clothing insulation levels of approximately 0.6
clo.

However, building occupants in winter typically wear substantially
different attire. Analysis of the ASHRAE Global Thermal Comfort Database
II [@foldvarylicinaDevelopmentASHRAEGlobal2018] indicates that median
clothing insulation in commercial spaces during heating seasons is
approximately 0.75 clo [@iso9920], with occupants commonly wearing long
trousers, closed-toe shoes and socks that cover the ankle region. This
discrepancy raises an important question: does the current ankle draft
model, which is calibrated for summer conditions and derived from
experiments where approximately two-thirds of test cases involved
exposed ankles, accurately predict discomfort when ankles are covered by
clothing?

If the current model overestimates draft risk under winter clothing
conditions, the practical implications are significant. Designers may be
specifying solutions, like perimeter heating systems, that are larger
than necessary, or installing them in situations where they could be
avoided entirely. This has consequences for both first costs and
operational energy use and related carbon emission. Conversely, if the
model underestimates risk, occupants may experience discomfort that
compromises satisfaction and performance.

The objective of this study is to evaluate thermal comfort responses to
ankle-level draft under typical winter clothing conditions through
controlled laboratory experiments. We aim to: (1) characterize
whole-body and local thermal perception across a range of ankle-level
air temperatures and air speeds representative of window downdraft
conditions; (2) compare observed dissatisfaction rates with predictions
from the existing ankle draft model; and (3) assess whether the current
ASHRAE 55 criteria require adjustment for winter scenarios. The findings
are intended to inform guidance on perimeter heating requirements and
support more energy-efficient building design without compromising
occupant comfort.

# Methods {#sec-methods}

## Experimental design

We conducted a controlled laboratory experiment using a within-subjects
repeated-measures design. Each participant was exposed to nine
experimental conditions in a 3×3 factorial arrangement, varying supply
air temperature (SAT) at three levels (15, 17, and 19°C) and air speed
at ankle level at three levels (low, medium, high; ranging between
0.1–0.7 m/s). The SAT and air speeds were selected to represent the
range of conditions associated with window downdraft under typical
winter conditions [@manzAnalysisThermalComfort2004] and to be consistent
with earlier studies focusing on summer conditions
[@liuPredictedPercentageDissatisfied2017;
@schiavonSensationDraftUncovered2016]

Participants attended four sessions: an introductory session for
explaining the experiment and obtaining consent, followed by three
experimental sessions of approximately two hours each. Each experimental
session corresponded to one SAT level, or ankle-level air temperature,
respectively, with participants experiencing all three air speed
conditions within that session in a counterbalanced order. We scheduled
sessions on separate days to minimize carryover effects and subjects
were allowed to attend the three sessions in their preferred order.

## Climate chamber and apparatus

We conducted the experiment in the climate-controlled chamber at the
Center for the Built Environment, UC Berkeley. The chamber floor area is
approximately 25 m². We set up three workstations to allow simultaneous
testing of multiple participants and three seats for reacclimatization
during breaks in the back of the room. Ambient room conditions were
maintained at approximately 22°C operative temperature, 50% relative
humidity, and background air velocity below 0.1 m/s, targeting a
whole-body thermal sensation of approximately −0.5 (slightly cool) on
the ASHRAE seven-point scale and as calculated using the CBE Comfort
Tool [@tartariniCBEThermalComfort2020].

![Experimental configuration and temporal protocol: (a) Seated occupant
positioned 0.7 m from the custom displacement diffuser, which simulates
window downdraft by directing conditioned air toward the ankle region
from behind; (b) Timeline of the 120-minute experimental sessions
featuring overlapping participant groups and synchronized questionnaire
intervals to characterize the progression of local and whole-body
thermal perception.](figs/experimental_setup.png){#fig-setup
width="75%"}

We used custom built displacement diffusers to deliver conditioned air
to the ankle region of participants seated at the three workstations. To
simulate a window downdraft, the diffuser was positioned 0.7 m behind
each workstation, directing airflow toward the participants' lower legs.
This setup targeted one of the most thermally sensitive body regions
highly susceptible to draft, as described by
@toftumEffectAirflowDirection1997. The diffuser was connected to an
independent air handling unit capable of delivering supply air at the
target temperatures and flow rates. The three supply air temperatures
represent temperatures at the VAV outlet and we controlled air flow
rates by damper adjustment. Turbulence intensity at the measurement
location was approximately 30% (see @tbl-env), consistent with values
reported in prior ankle draft studies
[@liuPredictedPercentageDissatisfied2017;
@fangerAirTurbulenceSensation1988]. The three seats in the back of the
room were not affected by the diffusers, which were directed at the
three workstations only.

## Monitoring

### Environmental

We measured head-level air temperature (1.1 m) at each workstation using
calibrated Atmocube monitors (Atmotech Inc., United States; air
temperature: -40 to +125°C, ±0.5°C, relative humidity: 0 to 100%RH, ±2
%RH, CO~2~: 0 to 5000 ppm, ±50 ppm + 2.5% of reading at standard room
levels) positioned within 0.5 m of participants. To measure air velocity
and air temperature at ankle level (0.1 m) we used the AirDistSys 5000
omnidirectional hot-wire anemometers (Sensor Electronic, Poland, air
velocity: 0.05-5 m/s, accuracy: ±0.02 m/s ±2% of reading; temperature:
-10 to +50°C, accuracy: ±0.2°C, sampling interval: 1 s). Room-level
measurements of CO~2~ concentration, PM~2.5~, illuminance and sound
pressure levels were recorded throughout each session but not reported
here because they stayed constant and within expected ranges.

### Physiological

In addition, we continuously recorded participants' skin temperature on
the lateral lower leg (mid-calf) and feet using iButton sensors
(Thermochron, Maxim Integrated, USA, Type DS1923; measurement range:
20°C to 85°C; accuracy: ±0.5°C; resolution: 0.0625°C) attached with
medical-grade hypoallergenic tape. These sites were selected as the most
practical iButton attachment locations in the lower extremity,
bracketing the ankle region; direct placement at the ankle was not
feasible due to the bony lateral malleolus. We used the skin temperature
data to assess local cooling responses and as physiological correlates
of subjective thermal sensation.

### Subjective feedback {#sec-subjective-feedback}

To assess the perception of the thermal environment we used validated
scales consistent with prior ankle draft research and ASHRAE Standard 55
requirements. We employed two questionnaires: a comprehensive one,
encompassing all questions, administered shortly before the conclusion
of the exposure (F), and a concise, partial one (P), comprising a subset
of the questions, at the end of each adaptation period. For details on
the timing of the questionnaires, please refer to @fig-setup, and for
information on the constructs and their respective scales, please refer
to @tbl-scales below. We administered the questions electronically via a
custom survey tool written in Java. Participants were not required to
respond to any individual question, though completion rates were
monitored.

```{r}
#| label: tbl-scales
#| tbl-cap: "Survey constructs and measurement scales, including information if the question was used in the full (F) and partial (P) subject feedback questionnaire."
#| tbl-colwidths: [15,10,25,15,35]
#| echo: false

# filtering for only the question we actually analyze
readr::read_csv("tables/survey_scales.csv") |>
  slice(1:4, 7:8, 10) |> 
  knitr::kable(align = "l")
```

## Participants

We recruited
`r demographic_d %>% dplyr::filter(sex == "All") %>% dplyr::pull(n)`
participants from the university community and local area through the
Experimental Social Science Laboratory (Xlab) platform and email
outreach to individuals who had participated in previous studies and
consented to future contact. Eligible participants were between 21 and
55 years of age, fluent in English, and free from chronic
cardiorespiratory conditions or recent major surgery. Pregnant
individuals were excluded. The UC Berkeley Committee for Protection of
Human Subjects (CPHS) approved the research protocol (CPHS
#2025-07-18795). Before the tests, all subjects signed an informed
consent form. Demographic details of the cohort are summarized in
@tbl-demographics.

## Experimental protocol

Upon arrival for each experimental session, we checked participants for
elevated temperature and fitted them with skin temperature sensors.
Participants then entered the climate chamber and completed a 20-minute
adaptation period in the acclimatization zone under baseline conditions.

To simulate typical winter attire, we instructed participants to wear a
long-sleeve top, long trousers (ending just above the ankle), thin
socks, and closed-toe shoes. This ensemble targets a clothing insulation
of approximately 0.75 clo, consistent with the median value observed in
commercial buildings during heating seasons
[@foldvarylicinaDevelopmentASHRAEGlobal2018]. Participants were
permitted to make minor clothing adjustments during sessions (e.g.,
rolling sleeves), and any changes were recorded.

Following adaptation, participants were exposed to three consecutive
20-minute conditions at different air speed levels, with 20-minute rest
periods between conditions (see @fig-setup). The 20-minute exposure
duration was selected to match the protocol used in the foundational
ankle draft studies [@liuPredictedPercentageDissatisfied2017;
@schiavonSensationDraftUncovered2016], enabling direct comparison of
results, and exceeds the 15-minute evaluation period specified in ASHRAE
Standard 55 for assessing local thermal discomfort
[@ashraeANSIASHRAEStandard2023]. The sequence of air speed conditions
was counterbalanced across participants and sessions. At the end of each
exposure and adaptation interval, participants completed the
standardized questionnaires listed in @sec-subjective-feedback.

Across the experiment, three participants completed six conditions and
another three completed three conditions for personal reasons. One
repeated trial resulting from an operational error was excluded, leaving
431 observations for analysis.

## Statistical analysis

### Variable processing

Thermal sensation votes were collected on a continuous scale. For
descriptive analyses, we mapped these values to the corresponding
categorical sensation levels, whereas the original continuous values
were retained for statistical modeling. We dichotomized ankle airflow
acceptability as acceptable or unacceptable for subsequent analyses. The
original recorded votes for thermal acceptability, thermal preference,
ankle thermal preference, and ankle airflow preference were retained
without further processing. Consistent with the definition used by
@liuPredictedPercentageDissatisfied2017, we defined ankle draft
dissatisfaction as an ankle thermal sensation vote below 0 (neutral)
together with an ankle airflow acceptability vote below 0. Draft is
defined as unwanted local cooling of the body caused by air movement
[@ashraeANSIASHRAEStandard2023]. Because acceptability and preference
capture different aspects of occupant response, we also conducted a
comparison analysis in the Appendix C using a preference-based
definition of ankle draft dissatisfaction.

### Descriptive regression

To visualise outcome–predictor relationships in the scatter plots, we
fitted per-group linear mixed-effects models with a random intercept for
subject to account for the within-subject dependence arising from the
repeated-measures design. We report the fixed-effect slope and the
marginal R², the proportion of variance explained by the fixed predictor
alone, following @nakagawa2013. Population-averaged prediction lines and
95% confidence intervals were derived analytically from the
fixed-effects variance–covariance matrix.

### Within-subject comparisons

We used a within-subject repeated-measures design and conducted
comparisons between experimental conditions within participants. We
analyzed continuous variables using paired-samples t-tests and reported
paired Cohen's $d_z$ as the effect size. We analyzed ordinal variables
using the paired Wilcoxon signed-rank test, reporting the test statistic
based on the normal approximation and the effect size $r$. For multiple
pairwise comparisons within each family of tests, we adjusted p-values
using the Benjamini-Hochberg method [@benjaminiHighberg1995].

### Evaluation of the existing ankle draft model

Applying the existing ankle draft model to the experimental data, we
calculated the predicted dissatisfaction rate for the sample at any
given condition.

We assessed model calibration by comparing predicted probabilities with
observed outcomes using a logistic recalibration model, following
@vancalster2016. The calibration curve indicates whether predicted
probabilities agree with observed proportions. For example, among
observations predicted to have a 30% dissatisfaction rate, about 30%
should actually be dissatisfied. The calibration intercept indicates
whether predictions are systematically too high or too low, and the
calibration slope indicates whether predictions are too extreme or not
extreme enough.

We evaluated model discrimination using the AUC (area under the ROC
curve). For example, an AUC of 0.8 in this study means that, if we
randomly select one dissatisfied person and one satisfied person, the
model will assign a higher predicted risk to the dissatisfied person
about 80% of the time. Following @hosmer_applied_2000, we interpreted
AUC values of 0.70–0.80, 0.80–0.90, and ≥0.90 as acceptable, excellent,
and outstanding discrimination, respectively. We considered values below
0.70 to indicate less-than-acceptable discrimination, with 0.50
representing no discrimination.

We assessed overall predictive accuracy using the Brier score
[@GLENNBrier1950]. For a binary outcome, the Brier score is the mean
squared difference between the predicted probability and the observed
outcome, where observed outcomes are ankle draft dissatisfaction (1) or
satisfaction (0). A Brier score of 0 is perfect and 1 is very poor.

$$
\mathrm{Brier\ score} = \frac{1}{N}\sum_{i=1}^{N}(p_i - y_i)^2,
$$

where $p_i$ is the predicted probability from the existing model for the
$i$-th observation, and $y_i$ is the corresponding binary outcome
indicating ankle draft dissatisfaction (1) or satisfaction (0).

In addition, predictions were averaged within each experimental
condition and compared with the corresponding observed percentages of
dissatisfaction to examine condition-specific prediction accuracy.

### Development of the new ankle draft model {#sec-method-new-model}

The existing ankle draft risk model was developed only for summer
conditions. To extend model applicability to broader boundary
conditions, we modeled ankle draft dissatisfaction using both the
dataset from this study and the dataset from
@liuPredictedPercentageDissatisfied2017. The sample sizes included in
the models are summarized in @tbl-model-datasets. Because both
experiments used repeated-measures designs, the observations were not
independent. We therefore used a generalized linear mixed model with a
binomial distribution and logit link.

```{r}
#| label: tbl-model-datasets
#| tbl-cap: "Datasets included in the development of the new ankle draft model."
#| echo: false

model_dataset_summary <- new_model_c %>%
  dplyr::mutate(
    dataset = dplyr::if_else(source == "toby", "Current study", "Liu et al. (2017)"),
    ankle_coverage = dplyr::if_else(clothing_type == "long", "Covered", "Uncovered")
  ) %>%
  dplyr::group_by(dataset, ankle_coverage) %>%
  dplyr::summarise(
    subjects = dplyr::n_distinct(subject_id),
    observations = dplyr::n(),
    .groups = "drop"
  ) %>%
  dplyr::arrange(factor(dataset, levels = c("Current study", "Liu et al. (2017)")),
                 factor(ankle_coverage, levels = c("Covered", "Uncovered")))

knitr::kable(
  model_dataset_summary,
  col.names = c("Dataset", "Ankle coverage status", "Subjects", "Observations")
)
```

Subject was included as a random intercept to account for
between-subject heterogeneity and within-subject dependence arising from
repeated measurements. Continuous predictors were mean-centered using
the overall means of the dataset before model fitting. Model performance
was evaluated using marginal and conditional $R^2$ [@nakagawa2013]. The
marginal $R^2$ quantifies the variance explained by the fixed effects,
whereas the conditional $R^2$ quantifies the variance explained by both
fixed and random effects.

Because the inverse-logit transformation is nonlinear, fixed-effect
predictions from a logistic mixed model do not directly represent
population-averaged probabilities. We therefore estimated marginal
predicted probabilities by integrating over the random-intercept
distribution using Monte Carlo simulation
[@neuhausKalbfleischHauck1991]. Because these marginal probabilities
generally cannot be expressed as a simple closed-form equation, we then
fitted logistic-form approximation to the Monte Carlo-simulated marginal
probabilities. The reported approximation $R^2$ values quantify the
agreement between the closed-form approximation and the Monte
Carlo-based marginal predictions. Details of this procedure are provided
in Appendix B.

### Sample size and statistical power

We conducted a simulation-based power analysis to assess whether our
sample of 51 participants was statistically adequate to detect the
primary effect of ankle air speed on draft dissatisfaction. At the
study-design stage, we used the existing short-duration experimental
dataset from @liuPredictedPercentageDissatisfied2017 as the best
available prior reference for a GLMM-based simulation sample-size
assessment. Specifically, we fitted a GLMM to the
@liuPredictedPercentageDissatisfied2017 dataset, which included 109
participants, and used this model as the reference model to estimate the
probability of detecting the fixed effect of ankle air speed across
different sample sizes. This analysis was intended to provide a
reference-scenario planning basis for the present sample size, rather
than a definitive estimate of the true statistical power for the winter
covered-ankle condition.

Using the `simr` package [@greenMacleod2016], we simulated 1,000
datasets at each of seven sample sizes ranging from 10 to 109
participants and estimated the proportion of simulations in which the
fixed effect of ankle air speed was detected at α = 0.05. The estimated
power reached 90.1% at N = 30 participants (95% CI: 88.1–91.9%),
exceeding the conventional 80% threshold. The estimated power for the
present sample size of N = 51 was 99.2% (95% CI: 98.4–99.7%), indicating
that the current sample size was sufficient to detect an air-speed
effect comparable to that observed in the
@liuPredictedPercentageDissatisfied2017 reference dataset.

## Software and data

We performed all data processing, statistical analyses, and model
fitting in R (version 4.4.1). The main packages we used were tidyverse
(version 2.0.0), lme4 (version 2.0.1), rstatix (version 0.7.3),
CalibrationCurves (version 3.0.0), and simr (version 1.0.6). Data and
analysis code are available on:
https://github.com/CenterForTheBuiltEnvironment/ankle-draft.

# Results {#sec-results}

## Participants and environmental conditions

```{r}
#| label: tbl-demographics
#| tbl-cap: "Summary of demographic information (median [IQR]) on study participants."

knitr::kable(demographic_d,
             col.names = c('Sex', 'No. Subjects', 'Age [a]', 'Height [m]', 'Weight [kg]', 'Body Mass Index (BMI)')
)

```

@tbl-demographics provides a summary of the demographic information
collected from the study participants. Since the demographic data was
not uniformly distributed, we use the median and interquartile range
(IQR) to describe the dataset. A total of
`r demographic_d %>% dplyr::filter(sex == "All") %>% dplyr::pull(n)`
subjects participated in the study
(`r demographic_d %>% dplyr::filter(sex == "Female") %>% dplyr::pull(n)`
female,
`r demographic_d %>% dplyr::filter(sex == "Male") %>% dplyr::pull(n)`
male), with a median age of
`r demographic_d %>% dplyr::filter(sex == "All") %>% dplyr::pull(age)`
years.

```{r}
#| label: tbl-env
#| tbl-cap: "Summary of environmental measurements."
knitr::kable(env_airflow_t,
             col.names = c('Measurement', '15°C', '17°C', '19°C')
)
```

@tbl-env provides an overview of the measured environmental conditions.
Across all three sessions, head-level air temperature remained stable at
`r env_airflow_t %>% dplyr::filter(measurement == "Air temperature, head level (°C)") %>% dplyr::pull("17-18°C")`
°C,
`r env_airflow_t %>% dplyr::filter(measurement == "Air temperature, head level (°C)") %>% dplyr::pull("18-19°C")`
°C, and
`r env_airflow_t %>% dplyr::filter(measurement == "Air temperature, head level (°C)") %>% dplyr::pull("19-20°C")`
°C, and relative humidity was comparable across sessions, confirming
stable background thermal conditions and establishing a baseline in line
with the PMV target level of \~ -0.5 (slightly cool).

As the primary experimental manipulation, we set supply air temperatures
to 15°C, 17°C, and 19°C to simulate the effect of window downdraft. As
shown in @tbl-env, ankle-level air temperatures, measured by the
hot-wire anemometer at 0.1 m, were consistently higher than the supply
setpoints due to mixing with the warmer ambient air, and reached ranges
of approximately 17–18°C, 18–19°C, and 19–20°C for the three conditions
respectively. Because the ankle-level temperature is the quantity
relevant to thermal comfort and to the draft risk model, all subsequent
analyses use the measured ankle-level air temperature rather than the
supply setpoint as the experimental grouping variable. Within each
session, we then exposed subjects to three airspeed levels at separate
workstations. We measured consistent air speeds across all three
workstations and sessions, confirming reliable reproduction of the low
(\~ 0.15 m/s), medium (\~ 0.4 m/s) and high (\~ 0.6 m/s) airspeed
conditions. Turbulence intensities remained low to medium throughout,
indicating stable and controlled airflow
[@fangerAirTurbulenceSensation1988].

## Whole body thermal perception

```{r}
#| label: whole-body-helpers
#| echo: false

# Helper functions for inline text
get_sensation_pct <- function(sat, ws, response) {
  thermal_sensation_overall_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

get_preference_pct <- function(sat, ws, response) {
  thermal_preference_overall_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

# Aggregated summaries for text
sensation_neutral_cool <- thermal_sensation_overall_summary %>%
  dplyr::filter(response_value %in% c("Neutral", "Slightly cool", "Cool")) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(pct = sum(pct), .groups = "drop")

# for ankle level comparison
sensation_slightlycool_cold <- thermal_sensation_overall_summary %>%
  dplyr::filter(response_value %in% c("Slightly cool", "Cool", "Cold")) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(pct = sum(pct), .groups = "drop")

preference_no_change <- thermal_preference_overall_summary %>%
  dplyr::filter(response_value == "No change")

preference_warmer <- thermal_preference_overall_summary %>%
  dplyr::filter(response_value == "Warmer")
```

@fig-thermal-sensation-overall shows whole-body thermal sensation votes
plotted against ankle air temperature and air speed. Votes ranged
predominantly from Cool to Slightly warm across conditions. Thermal
sensation increased with ankle-level air temperature in all air speed
groups, with steeper slopes at medium and high air speeds (b ≈
`r round(min(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_overall", x_var == "t_supply_c", group %in% c("medium", "high")) |> dplyr::pull(slope)), 2)`
–
`r round(max(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_overall", x_var == "t_supply_c", group %in% c("medium", "high")) |> dplyr::pull(slope)), 2)`
per °C) than at low air speed (b ≈
`r round(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_overall", x_var == "t_supply_c", group == "low") |> dplyr::pull(slope), 2)`
per °C); however, ankle-level air temperature explained no variance
[@fergusonEffectSizePrimer2009] across all groups (R² ≤
`r round(max(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_overall", x_var == "t_supply_c") |> dplyr::pull(r2)), 2)`).
No systematic relationship was observed between whole-body thermal
sensation and ankle air speed, consistent with ankle-level airflow
having limited influence on whole-body thermal perception (see Appendix
A for response distributions).

![Whole-body thermal sensation votes plotted against ankle air
temperature (a) and air speed (b), with per-group linear mixed-effects
regression lines and 95% confidence intervals. Sensations were
predominantly in the Cool to Slightly warm range; ankle-level air
temperature showed a modest positive association at medium and high air
speeds, while no clear relationship with air speed was
observed.](figs/thermal_sensation_overall.png){#fig-thermal-sensation-overall
width="100%"}

```{r}
#| label: acceptability-helpers
#| echo: false

# Thermal acceptability summary
thermal_acceptability_summary <- thermal_acceptability_d %>%
  dplyr::mutate(acceptable = response_value > 0) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(
    n = dplyr::n(),
    pct_acceptable = mean(acceptable),
    mean_value = mean(response_value),
    .groups = "drop"
  )

acceptability_range <- thermal_acceptability_summary %>%
  dplyr::summarise(
    min = scales::percent(min(pct_acceptable), accuracy = 1),
    max = scales::percent(max(pct_acceptable), accuracy = 1)
  )

acceptability_by_sat <- thermal_acceptability_summary %>%
  dplyr::group_by(session_sat) %>%
  dplyr::summarise(
    min = scales::percent(min(pct_acceptable), accuracy = 1),
    max = scales::percent(max(pct_acceptable), accuracy = 1),
    .groups = "drop"
  )
```

@fig-thermal-acceptability-overall shows whole-body thermal
acceptability plotted against ankle air temperature and air speed.
Responses were predominantly on the acceptable side of the scale across
all conditions. No practically relevant linear relationship was found
between thermal acceptability and either ankle-level air temperature or
air speed (R² ≤
`r round(max(lme_stats_d |> dplyr::filter(outcome == "thermal_acceptability_overall") |> dplyr::pull(r2)), 3)`
in all groups), indicating that the range of conditions tested did not
substantially compromise whole-body thermal comfort (see Appendix A for
response distributions).

![Whole-body thermal acceptability votes plotted against ankle air
temperature (a) and air speed (b), with per-group linear mixed-effects
regression lines and 95% confidence intervals. Acceptability was
predominantly positive across all conditions, with no strong linear
association with ankle-level air temperature or air
speed.](figs/thermal_acceptability_overall.png){#fig-thermal-acceptability-overall
width="100%"}

## Ankle level observations

### Thermal perception

```{r}
#| label: ankle-helpers
#| echo: false

# Helper functions for ankle-level inline text
get_ankle_sensation_pct <- function(sat, ws, response) {
  thermal_sensation_ankle_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

get_ankle_preference_pct <- function(sat, ws, response) {
  thermal_preference_ankle_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

# Aggregated ankle sensation on cool side (cold + cool + slightly cool)
ankle_sensation_cool_side <- thermal_sensation_ankle_summary %>%
  dplyr::filter(response_value %in% c("Cold", "Cool", "Slightly cool")) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(pct = sum(pct), .groups = "drop")

# Ankle preference for warmer
ankle_preference_warmer <- thermal_preference_ankle_summary %>%
  dplyr::filter(response_value == "Warmer")
```

@fig-thermal-sensation-ankle presents ankle-level thermal sensation
votes plotted against ankle air temperature and air speed. Sensations at
the ankles were generally cooler than whole-body ratings. Ankle-level
air temperature showed a positive association with ankle sensation
across all air speed groups (b ≈
`r round(min(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "t_supply_c") |> dplyr::pull(slope)), 2)`
–
`r round(max(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "t_supply_c") |> dplyr::pull(slope)), 2)`
per °C), consistent with whole-body results. The relationship with ankle
air speed was negative across all ankle-level air temperature conditions
and strongest at 17–18°C (b =
`r round(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group == "17-18°C") |> dplyr::pull(slope), 2)`,
R² =
`r round(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group == "17-18°C") |> dplyr::pull(r2), 2)`),
indicating that higher air speeds increased local cool sensations most
noticeably at the coolest ankle-level air temperature. The association
was weaker at 18–19°C and 19–20°C (b ≈
`r round(max(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group %in% c("18-19°C", "19-20°C")) |> dplyr::pull(slope)), 2)`
to
`r round(min(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group %in% c("18-19°C", "19-20°C")) |> dplyr::pull(slope)), 2)`,
R² ≤
`r round(max(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group %in% c("18-19°C", "19-20°C")) |> dplyr::pull(r2)), 2)`)
(see Appendix A for response distributions).

![Ankle-level thermal sensation votes plotted against ankle air
temperature (a) and air speed (b), with per-group linear mixed-effects
regression lines and 95% confidence intervals. Ankle sensations were
generally cooler than whole-body ratings; the strongest negative
association with air speed was observed at the lowest ankle-level air
temperature condition
(17–18°C).](figs/thermal_sensation_ankle.png){#fig-thermal-sensation-ankle
width="100%"}

```{r}
#| label: air-movement-helpers
#| echo: false

# Air movement acceptability summary
air_acceptability_acceptable <- air_movement_acceptability_summary %>%
  dplyr::filter(acceptability == "Acceptable")

air_acceptability_by_sat <- air_acceptability_acceptable %>%
  dplyr::group_by(session_sat) %>%
  dplyr::summarise(
    min = scales::percent(min(pct), accuracy = 1),
    max = scales::percent(max(pct), accuracy = 1),
    .groups = "drop"
  )

# Air movement preference for less
air_pref_less <- air_movement_preference_summary %>%
  dplyr::filter(acceptability == "Less")

air_pref_no_change <- air_movement_preference_summary %>%
  dplyr::filter(acceptability == "No change")
```

### Air movement perception

@fig-air-move-acceptability shows air movement acceptability at the
ankle level plotted against ankle-level air temperature and air speed.
Despite the localized cooling, responses were predominantly on the
acceptable side of the scale. Air movement acceptability showed a
consistent negative association with air speed across all ankle-level
air temperature conditions (b ≈
`r round(max(lme_stats_d |> dplyr::filter(outcome == "air_movement_acceptability", x_var == "v_air_m_s") |> dplyr::pull(slope)), 2)`
to
`r round(min(lme_stats_d |> dplyr::filter(outcome == "air_movement_acceptability", x_var == "v_air_m_s") |> dplyr::pull(slope)), 2)`
per m/s), while the relationship with ankle-level air temperature was
negligible for the low and high air speed groups (b ≈
`r round(mean(lme_stats_d |> dplyr::filter(outcome == "air_movement_acceptability", x_var == "t_supply_c", group %in% c("low", "high")) |> dplyr::pull(slope)), 2)`)
and modest for the medium group (b ≈
`r round(lme_stats_d |> dplyr::filter(outcome == "air_movement_acceptability", x_var == "t_supply_c", group == "medium") |> dplyr::pull(slope), 2)`).
Linear regression explained a negligible [@fergusonEffectSizePrimer2009]
proportion of variance in all groups (R² ≤
`r round(max(lme_stats_d |> dplyr::filter(outcome == "air_movement_acceptability") |> dplyr::pull(r2)), 3)`),
indicating substantial individual variability in acceptability ratings
(see Appendix A for response distributions).

![Ankle-level air movement acceptability votes plotted against ankle air
temperature (a) and air speed (b), with per-group linear mixed-effects
regression lines and 95% confidence intervals. Acceptability was
predominantly positive; a consistent negative association with air speed
was observed across all ankle-level air temperature
conditions.](figs/air_movement_acceptability_ankles.png){#fig-air-move-acceptability
width="100%"}

### Skin temperature

```{r}
#| label: tsk-delta-helpers
#| echo: false

tsk_delta_last1_range <- tsk_delta_last1 %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(mean_delta_last1_c = mean(mean_delta_last1_c, na.rm = TRUE), .groups = "drop") %>%
  dplyr::summarise(
    min_val = round(min(mean_delta_last1_c, na.rm = TRUE), 1),
    max_val = round(max(mean_delta_last1_c, na.rm = TRUE), 1)
  )

tsk_delta_sig_low_med_all <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "medium") %>%
  dplyr::summarise(all_sig = all(p.adj < 0.05, na.rm = TRUE)) %>%
  dplyr::pull(all_sig)

tsk_delta_sig_low_med_max_p <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "medium") %>%
  dplyr::summarise(max_p = max(p.adj, na.rm = TRUE)) %>%
  dplyr::pull(max_p)

tsk_delta_sig_low_high_all <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "high") %>%
  dplyr::summarise(all_sig = all(p.adj < 0.05, na.rm = TRUE)) %>%
  dplyr::pull(all_sig)

tsk_delta_sig_low_high_max_p <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "high") %>%
  dplyr::summarise(max_p = max(p.adj, na.rm = TRUE)) %>%
  dplyr::pull(max_p)

stopifnot(isTRUE(tsk_delta_sig_low_med_all))
stopifnot(isTRUE(tsk_delta_sig_low_high_all))

tsk_delta_sig_med_high <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "medium", group2 == "high") %>%
  dplyr::mutate(sig = p.adj < 0.05) %>%
  dplyr::filter(sig) %>%
  dplyr::pull(session_sat)

tsk_delta_sig_med_high_p <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "medium", group2 == "high", p.adj < 0.05) %>%
  dplyr::pull(p.adj)

tsk_delta_low_vs_higher_d_range <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 %in% c("medium", "high"), p.adj < 0.05) %>%
  dplyr::summarise(
    min_val = sprintf("%.1f", min(abs(cohen_d), na.rm = TRUE)),
    max_val = sprintf("%.1f", max(abs(cohen_d), na.rm = TRUE))
  )

format_p_threshold <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "p_adj < 0.001",
    p < 0.01 ~ "p_adj < 0.01",
    p < 0.05 ~ "p_adj < 0.05",
    TRUE ~ sprintf("p_adj = %.3f", p)
  )
}
```

@fig-tsk illustrates the progression of lower-leg skin temperatures
throughout the exposure period. Using the pre-exposure adaptation period
as a baseline, lower-leg skin temperatures decreased under all
conditions, with final changes ranging from
`r tsk_delta_last1_range$min_val`°C to
`r tsk_delta_last1_range$max_val`°C. In general, the magnitude of this
skin temperature reduction increased with air speed, as medium and high
air speeds produced greater local cooling than low air speed across all
three ankle-level air temperatures.

Statistical analysis with Benjamini-Hochberg correction showed that the
differences in cooling response between low and medium air speeds, and
between low and high air speeds, were significant at all ankle-level air
temperatures, with maximum adjusted p-values of
`r format_p_threshold(tsk_delta_sig_low_med_max_p)` and
`r format_p_threshold(tsk_delta_sig_low_high_max_p)`, respectively. In
contrast, the difference between medium and high air speeds reached
significance only at `r tsk_delta_sig_med_high`
(`r sprintf("p_adj = %.3f", tsk_delta_sig_med_high_p)`). These findings
indicate that increasing local airflow increases the short-term cooling
response of the lower-leg skin and most noticeable temperature drops
occur when shifting away from the lowest air speed.

![Time course of lower-leg skin temperature during the formal exposure
phase across the nine experimental conditions for (a) absolute skin
temperature and (b) change relative to the adaptation-period baseline (
Shaded bands denote the between-subject mean ± 1 SD): Local cooling
intensified as air speed increased, with the most significant
temperature drops occurring when moving from low to medium
airflow.](figs/tsk_p.png){#fig-tsk width="90%"}

```{r}
#| label: tsk-extrapolation
#| echo: false

tsk_delta_extrapolation_range <- tsk_delta_exponential_fits %>%
  dplyr::summarise(
    min_val = sprintf("%.1f", min(abs(additional_delta_20_to_90), na.rm = TRUE)),
    max_val = sprintf("%.1f", max(abs(additional_delta_20_to_90), na.rm = TRUE))
  )
```

It should be noted that lower-leg skin temperature had not fully
stabilized by the end of the 20 min exposure. To estimate the potential
effect of exposure duration on the results, we further fitted an
exponential regression model to the measured skin temperature changes
during the first 20 min and extrapolated the fitted trend from 20 to 90
min, as shown in @fig-tsk-delta-extrapolation. The results indicate that
lower-leg skin temperature would continue to decrease after 20 min, with
an additional reduction of approximately
`r tsk_delta_extrapolation_range$min_val`–`r tsk_delta_extrapolation_range$max_val`
°C between 20 and 90 min before stabilizing. This extrapolation was
broadly consistent with simulations from the Advanced Berkeley Comfort
(ABC) model [@ABCHuizenga2001] under the same experimental conditions,
which predicted an additional lower-leg skin temperature decrease of
approximately 0.3–0.7 °C. The corresponding changes in both local and
whole-body thermal sensation were below 0.3 on the TSV scale predicted
by ABC model. Therefore, although lower-leg skin temperature had not
fully stabilized, the projected additional cooling is modest and the
main conclusions are unlikely to be substantially altered by longer
exposure durations. Nevertheless, the effect on subjective
dissatisfaction beyond 20 minutes cannot be directly assessed from the
current data, and dissatisfaction rates under prolonged real-office
exposure may be somewhat higher than those reported here.

![Estimated lower-leg skin temperature changes across experimental
conditions. Solid lines show measured mean values during the first 20
min, and dashed lines show exponential extrapolations from 20 to 90
min.](figs/tsk_delta_extrapolation_p.png){#fig-tsk-delta-extrapolation
width="90%"}

## Dissatisfaction with ankle draft {#sec-model-comp}

```{r}
#| label: setup-results-data
#| echo: false
#| message: false

# Converting decimals to percentages and rounding for cleaner inline text
res <- model_comparison_d %>%
  mutate(across(c(ppd_liu_mean, dissatisfied_rate, diff), ~ round(.x * 100, 1)))

liu_model_summary <- model_comparison_d %>%
  dplyr::summarise(
    mean_observed = round(mean(dissatisfied_rate) * 100, 0),
    mean_predicted = round(mean(ppd_liu_mean) * 100, 0),
    overprediction_ratio = round(mean(ppd_liu_mean) / mean(dissatisfied_rate), 1)
  )

liu_diff_by_speed <- model_comparison_d %>%
  dplyr::group_by(session_air_flow) %>%
  dplyr::summarise(mean_diff = mean(diff), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_diff))

liu_max_diff_speed <- liu_diff_by_speed %>% dplyr::slice(1) %>% dplyr::pull(session_air_flow)
liu_min_diff_speed <- liu_diff_by_speed %>% dplyr::slice(n()) %>% dplyr::pull(session_air_flow)
```

<!-- FLAG: hardcoded "15°C", "17°C", "19°C" ankle temperature references updated to measured ranges; verify these index positions (res$diff[1], [3], [9]) still align with the correct conditions after any data reordering -->

For each observation, we calculated the predicted percentage of
dissatisfaction with ankle draft using the existing ankle draft model
based on ankle-level air speed and actual whole-body thermal sensation.
@fig-model-comp(a) shows the resulting calibration curve. In a
well-calibrated model, the calibration curve should closely follow the
45° ideal line. Although the existing model had a Brier score of 0.14,
the systematic deviation of the calibration curve from the ideal line
indicates that the existing model consistently overestimated the
probability of ankle draft dissatisfaction under winter conditions. In
addition, the existing model retained acceptable discriminative ability
in the present dataset (AUC =
`r sprintf("%.2f", liu_calibration_stats["C (ROC)"])`)
[@hosmer_applied_2000].

@fig-model-comp(b) compares the dissatisfaction rates predicted by the
existing model against our experimental observations under each
experimental condition. Across all conditions, the existing model
consistently overpredicted local thermal dissatisfaction. At a 17–18°C
ankle-level air temperature, the discrepancy between predicted and
observed rates widened as air speed increased: the model overpredicted
dissatisfaction by `r res$diff[1]`%, rising to `r res$diff[3]`% at high
flow (0.6 m/s). We observed similar trends at 18–19°C and 19–20°C.
Notably, for the 19–20°C at high air speed (0.6 m/s), the model
predicted a PPD of `r res$ppd_liu_mean[9]`%, while the actual
dissatisfied rate was only `r res$dissatisfied_rate[9]`%, resulting in a
mean overestimation of `r res$diff[9]`%. Across the nine experimental
conditions, the mean observed dissatisfaction rate was
`r liu_model_summary$mean_observed`%, whereas the Liu model predicted an
average of `r liu_model_summary$mean_predicted`%, corresponding to an
overestimation of approximately
`r liu_model_summary$overprediction_ratio` times.

![Performance of the existing ankle draft model: (a) calibration curve
of the existing model on this study dataset, and (b) predicted and
observed dissatisfaction under each experimental condition (points
represent mean values, with the segment length indicating the magnitude
of model
overestimation)](figs/model_performance_combined_raw.png){#fig-model-comp
width="100%"}

## Updated model for winter conditions

```{r}
#| label: setup-updated-model
#| echo: false
#| message: false

model_input_coef <- summary(model_input)$coefficients

model_screening <- tibble::tibble(
  term = rownames(model_input_coef),
  estimate = model_input_coef[, "Estimate"],
  std_error = model_input_coef[, "Std. Error"],
  z_value = model_input_coef[, "z value"],
  p_value = model_input_coef[, "Pr(>|z|)"]
)

model_screening_table <- model_screening %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::mutate(
    term_label = dplyr::recode(
      term,
      "v_air_m_s" = "Air speed at ankle",
      "sexmale" = "Sex (male vs. female [reference])",
      "t_air_c" = "Room air temperature",
      "t_supply_c" = "Air temperature at ankle",
      "thermal_sensation" = "Whole-body thermal sensation",
      "turbulence_intensity" = "Turbulence intensity",
      "clothing_typeshort" = "Ankle coverage status (uncovered vs. covered [reference])",
      "sourcetoby" = "Data source (This study vs. Liu's summer dataset [reference])"
    ),
    estimate_fmt = sprintf("%.2f", estimate),
    std_error_fmt = sprintf("%.2f", std_error),
    z_value_fmt = sprintf("%.2f", z_value),
    p_value_fmt = dplyr::case_when(
      p_value < 0.001 ~ "<0.001",
      TRUE ~ sprintf("%.3f", p_value)
    )
  ) %>%
  dplyr::select(term_label, estimate_fmt, std_error_fmt, z_value_fmt, p_value_fmt)

model_formula_inline <- model_formula %>%
  dplyr::mutate(
    intercept_fmt = sprintf("%.3f", Intercept),
    v_fmt = sprintf("%.3f", V_coef),
    r2_fmt = sprintf("%.4f", R2),
    formula_text_expanded = sprintf(
      r"(PPD_{AD} = \frac{\exp\left(%s + %sV %s %sTS\right)}{1 + \exp\left(%s + %sV %s %sTS\right)})",
      intercept_fmt,
      v_fmt,
      ifelse(TS_coef < 0, "-", "+"),
      sprintf("%.3f", abs(TS_coef)),
      intercept_fmt,
      v_fmt,
      ifelse(TS_coef < 0, "-", "+"),
      sprintf("%.3f", abs(TS_coef))
    )
  )

model_exposed <- model_formula_inline %>% dplyr::filter(Model == "exposed")
model_covered <- model_formula_inline %>% dplyr::filter(Model == "unexposed")

glmm_exposed_coef <- lme4::fixef(m_glmm_exposed)
glmm_covered_coef <- lme4::fixef(m_glmm_unexposed)
glmm_delta_intercept <- glmm_covered_coef["(Intercept)"] - glmm_exposed_coef["(Intercept)"]

extract_final_model_table <- function(mod, model_label) {
  coef_mat <- summary(mod)$coefficients
  tibble::tibble(
    model = model_label,
    term = rownames(coef_mat),
    estimate = coef_mat[, "Estimate"],
    std_error = coef_mat[, "Std. Error"],
    z_value = coef_mat[, "z value"],
    p_value = coef_mat[, "Pr(>|z|)"]
  )
}

extract_final_model_r2 <- function(mod, model_label) {
  r2 <- performance::r2_nakagawa(mod)
  tibble::tibble(
    model = model_label,
    observations = stats::nobs(mod),
    marginal_r2 = sprintf("%.3f", r2$R2_marginal),
    conditional_r2 = sprintf("%.3f", r2$R2_conditional)
  )
}

model_final_r2 <- dplyr::bind_rows(
  extract_final_model_r2(m_glmm_exposed, "Ankle uncovered"),
  extract_final_model_r2(m_glmm_unexposed, "Ankle covered")
)

model_final_table <- dplyr::bind_rows(
  extract_final_model_table(m_glmm_exposed, "Ankle uncovered"),
  extract_final_model_table(m_glmm_unexposed, "Ankle covered")
) %>%
  dplyr::left_join(model_final_r2, by = "model") %>%
  dplyr::mutate(
    term_label = dplyr::recode(
      term,
      "(Intercept)" = "Intercept",
      "v_air_m_s" = "Air speed at ankle",
      "thermal_sensation" = "Whole-body thermal sensation"
    ),
    estimate_fmt = sprintf("%.2f", estimate),
    std_error_fmt = sprintf("%.2f", std_error),
    z_value_fmt = sprintf("%.2f", z_value),
    p_value_fmt = dplyr::case_when(
      p_value < 0.001 ~ "<0.001",
      TRUE ~ sprintf("%.3f", p_value)
    )
  ) %>%
  dplyr::group_by(model) %>%
  dplyr::mutate(
    model_display = dplyr::if_else(dplyr::row_number() == 1, model, ""),
    observations_display = dplyr::if_else(
      dplyr::row_number() == 1, as.character(observations), ""
    ),
    marginal_r2_display = dplyr::if_else(dplyr::row_number() == 1, marginal_r2, ""),
    conditional_r2_display = dplyr::if_else(dplyr::row_number() == 1, conditional_r2, "")
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(
    model_display, observations_display, term_label, estimate_fmt,
    std_error_fmt, z_value_fmt, p_value_fmt, marginal_r2_display,
    conditional_r2_display
  )
```

As shown in @sec-model-comp, the existing model consistently
overestimated ankle draft dissatisfaction under winter conditions. This
highlights the need for a revised model that accounts for differences in
application contexts. To enable an integrated analysis of the effects of
different factors on ankle draft dissatisfaction, we combined data from
@liuPredictedPercentageDissatisfied2017 and collected in this study in
the modeling dataset. The combined dataset is available at
<https://github.com/CenterForTheBuiltEnvironment/ankle-draft>.

To identify significant predictors, we selected candidate variables
based on the initial variable screening procedure used by
@liuPredictedPercentageDissatisfied2017. These included environmental
factors, including ankle-level air speed, ankle-level air temperature,
room air temperature, and ankle-level turbulence intensity. Based on the
existing ankle draft risk model, we also included sex and whole-body
thermal sensation. In addition, we included ankle coverage status to
examine clothing influence, and data source to examine potential
systematic differences between the datasets after controlling for the
other predictors.

We first fitted a model including all candidate predictors to the
combined dataset observations (n =
`r format(nobs(model_input), big.mark = ",")`), following the methods
described in @sec-method-new-model. The result of the screening model is
shown in @tbl-updated-model-screening. Ankle-level air speed, whole-body
thermal sensation, and ankle coverage status (uncovered vs. covered)
were significantly correlated with ankle draft dissatisfaction, whereas
room air temperature, ankle-level air temperature, turbulence intensity,
and sex were not significant within the range investigated in this
study. The data source was also not statistically significant,
suggesting that no significant systematic difference remained between
the two datasets after accounting for the measured predictors.

```{r}
#| label: tbl-updated-model-screening
#| tbl-cap: "Results of the generalized linear mixed-effects model used for predictor screening."
#| echo: false

knitr::kable(
  model_screening_table,
  col.names = c("Predictor", "Estimate", "SE", "z", "p")
)
```

Based on these results, we developed separate mixed-effects models for
ankle draft dissatisfaction under the ankle-uncovered and ankle-covered
conditions, using ankle-level air speed and whole-body thermal sensation
as predictors. As shown in @tbl-updated-model-final, the two final
models had the same structure but different coefficients. The intercept
of the covered model was `r sprintf("%.1f", abs(glmm_delta_intercept))`
units lower than that of the uncovered model, indicating a lower
baseline risk of ankle draft dissatisfaction when the ankle was covered
by clothing. In addition, the air speed coefficient was smaller in the
covered model than in the uncovered model, indicating reduced
sensitivity of dissatisfaction to air speed under ankle-covered
conditions.

```{r}
#| label: tbl-updated-model-final
#| tbl-cap: "Generalized linear mixed-effects models for the ankle-uncovered and ankle-covered conditions."
#| echo: false

knitr::kable(
  model_final_table,
  col.names = c(
    "Condition", "Observations",  "Predictor", "Estimate", "SE", "z", "p", "Marginal R²", "Conditional R²"
  )
)
```

As described in @sec-method-new-model, although the conditional
prediction from the mixed-effects model has a simple logistic form, the
population-averaged marginal prediction requires integration over the
random-intercept distribution and generally has no simple closed-form
expression. Therefore, we used Monte Carlo simulation to obtain the
marginal predictions and fitted logistic-form approximations to them.
The resulting models are shown in @fig-updated-model.

![Updated ankle draft dissatisfaction model under exposed and covered
conditions. (a) ankle-uncovered conditions, and (b) ankle-covered
conditions.](figs/Model.png){#fig-updated-model width="100%"}

For the ankle-uncovered condition, the corresponding approximate
closed-form equation (approximation $R^2$ = `r model_exposed$r2_fmt`)
is:

```{r}
#| label: formula-model-exposed
#| echo: false
#| results: asis
cat("$$\n", model_exposed$formula_text_expanded, "\n$$\n", sep = "")
```

For the ankle-covered condition, the corresponding approximate
closed-form equation (approximation $R^2$ = `r model_covered$r2_fmt`)
is:

```{r}
#| label: formula-model-covered
#| echo: false
#| results: asis
cat("$$\n", model_covered$formula_text_expanded, "\n$$\n", sep = "")
```

Where, $PPD_{AD}$ is the predicted percentage dissatisfied with ankle
draft, $V$ is air speed at ankle (m/s), and $TS$ is whole-body thermal
sensation. Consistent with the PPD framework established by
@fangerThermalComfortAnalysis1970 and adopted in ASHRAE Standard 55, the
models are intended to predict population-average responses, or the
expected proportion of occupants dissatisfied under given conditions,
rather than the thermal response of any individual occupant. The
substantial inter-individual variability evident in Figures 3–6 is
expected and is accounted for in the mixed-effects model structure
through the random subject intercept. In addition, the proposed model
was developed based on the experimental ranges covered by the present
study and Liu‘s study. Therefore, its application should account for the
range of experimental conditions represented in these datasets to avoid
inappropriate extrapolation.

# Discussion

## Comparison with prior research

The central finding of this study is that ankle-level dissatisfaction
under winter clothing conditions (≥0.75 clo) is substantially lower than
predicted by the existing model of
@liuPredictedPercentageDissatisfied2017 and than observed in prior
studies with exposed ankles. In the experiments of
@schiavonSensationDraftUncovered2016, where participants wore summer
clothing (\~0.6 clo) with exposed ankles and lower legs, unacceptability
rates ranged from 23% to 57%. Under the ankle-covered conditions in this
study, airflow unacceptability ranged from only 6% to 20% across
comparable air speed and temperature conditions.

These findings align with other research on the protective effects of
lower-body coverage. @zhouDraughtSensationAssessment2022 reported that
high-top shoes increased ankle airflow acceptability by approximately
15% in stratified environments, demonstrating that even partial ankle
coverage meaningfully reduces sensitivity to local airflow. Similarly,
@toftumDraughtSensitivityInfluenced1996 found that occupants with cooler
overall thermal sensations are more likely to perceive airflow as
uncomfortable, and that higher clothing insulation attenuates the
interaction between ambient temperature and airflow-induced discomfort.

The limited variance explained by the fixed-effect predictors in the
mixed-effects regression models shown in @fig-thermal-sensation-ankle
and @fig-air-move-acceptability further supports this interpretation.
Across all conditions, R² values did not exceed
`r round(max(lme_stats_d$r2), 2)`, with the strongest association
observed for ankle-level thermal sensation versus air speed at 17–18°C
(R² =
`r round(lme_stats_d |> dplyr::filter(outcome == "thermal_sensation_ankles", x_var == "v_air_m_s", group == "17-18°C") |> dplyr::pull(r2), 2)`).
@liuPredictedPercentageDissatisfied2017 reported an R² of 0.10 for ankle
air movement acceptability versus air speed alone, and an adjusted R² of
0.22 for a model including both whole-body thermal sensation and ankle
air speed, substantially higher than our values and consistent with the
greater sensitivity to ankle draft expected under summer clothing
conditions with exposed ankles.

Regarding predictor selection, only whole-body thermal sensation and air
speed were significant predictors of dissatisfaction in our models,
consistent with the structures reported by
@toftumDraughtSensitivityInfluenced1996 and
@liuPredictedPercentageDissatisfied2017.
@chengExperimentalStudyDraftrelated2023 similarly identified a
relationship between overall thermal sensation and draft rate
thresholds, further supporting its inclusion as a predictor.

Although the dataset used by @liuPredictedPercentageDissatisfied2017
included both exposed and covered ankle conditions, no significant
clothing effect was identified in their analysis even if the ankle
condition (covered vs uncovered) was the third most important variable
after ankle air speed and whole-body thermal sensation. However, when we
reanalyzed their dataset using a mixed-effects model structure, the
clothing effect became statistically significant (p = 0.00423). The
discrepancy in the significance of this variable likely stems from their
use of standard logistic regression, which pooled observations across
subjects without accounting for the non-independence introduced by
repeated measures. This approach may have masked the clothing effect as
a between-subject variable due to inter-individual variability, which is
a limitation addressed in our study through the use of mixed-effects
models.

## Mechanisms

Two complementary pathways, physiological and psychological, may explain
the reduced sensitivity to ankle draft under winter clothing conditions.

From a physiological perspective, the skin temperature results indicate
that winter clothing attenuates, though does not eliminate, the local
cooling effect of airflow. During the final steady-state period,
significant reductions in lower-leg skin temperature occurred only at
the coolest ankle-level air temperature condition (17–18°C), where
medium and high air speeds produced lower skin temperatures than the
low-speed condition. At 18–19°C and 19–20°C, variations in air speed did
not yield significant differences in steady-state skin temperature, and
no significant differences were observed between medium and high air
speeds at any ankle-level air temperature condition. These findings
suggest that detectable local cooling requires a sufficiently large
environmental contrast.

Thermal sensation changes are more closely associated with rapid
variations in skin temperature than with absolute values
[@ChoiInvestigationHumanBody2012; @KenshaloCoolThresholds1968]. In this
study, ankle skin temperature variation relative to the baseline
adaptation period showed significant differences between low and
medium/high air speed conditions across all ankle-level air temperature
conditions (effect size *d* ≈
`r tsk_delta_low_vs_higher_d_range$min_val`–`r tsk_delta_low_vs_higher_d_range$max_val`).
Consistent with this, local ankle thermal sensation at medium and high
speeds was significantly lower than at low speed under the 17–18°C
condition. However, air movement acceptability did not differ
significantly across conditions, indicating that while air speed
produced measurable shifts in skin temperature and local thermal
sensation, the magnitude of these changes was insufficient to generate
distinguishable differences in dissatisfaction within the tested range.
From a heat transfer perspective, clothing coverage increases local
thermal resistance and reduces convective and radiative heat exchange
[@oguroEvaluationEffectAir2001; @nishiDirectMeasurementClothing1975]. It
should be noted that the clothing insulation of 0.75 clo in this study
represents the whole-body ensemble, whereas the local insulation at the
ankle was provided by thin socks (approximately 0.02 clo
@ashraeANSIASHRAEStandard2023). Despite this low local insulation, we
observed that ankle coverage still suppressed both the mxagnitude and
variability of skin temperature changes and made it less likely that
occupants cross discomfort thresholds.

Psychological factors may also contribute to the observed tolerance.
Occupants wearing thicker lower-body clothing may perceive themselves as
"protected," increasing their tolerance to local airflow. Prior research
has demonstrated that thermal perception is not solely determined by
physical stimuli but is modulated by expectations and informational
cues. Providing occupants with temperature information or group
perception feedback can improve thermal comfort evaluations
[@zhangGroupPerceptionFeedback2022], and thermal expectations themselves
influence perception [@schweikerThermalExpectation2020]. Field studies
have also identified associations between perceived control and thermal
satisfaction in winter conditions [@xuPerceivedControlWinter2021]. These
findings suggest that the subjective sense of being protected by
clothing may enhance tolerance to ankle draft, though this psychological
pathway was not directly measured in the present study and warrants
further investigation.

Together, these physiological and psychological mechanisms likely act in
concert: clothing reduces the physical cooling stimulus while
simultaneously altering occupants' expectations and perceived
vulnerability to draft.

## Implications for building design and standards

The ankle draft model of @liuPredictedPercentageDissatisfied2017, now
incorporated into ASHRAE Standard 55, does not distinguish between
covered and uncovered ankle conditions. Given the standard's acceptable
thermal sensation range of PMV = −0.5 to +0.5, the current criterion
requires ankle-level air speed below 0.22 m/s to achieve less than 20%
dissatisfied when TS = −0.5. This constraint may be overly conservative
for typical winter environments where occupants wear clothing that
covers the ankle.

The findings of this study support a differentiated approach. For winter
conditions with ankle coverage (≥0.75 clo), the covered-ankle model
developed here indicates that dissatisfaction remains within acceptable
limits at air speeds that would exceed the current threshold. Empirical
predictions of window downdraft suggest that maximum velocities at
approximately 1 m from the façade are around 0.25 m/s, with peak
velocities often falling within 0.1–0.2 m/s under milder outdoor
conditions or with smaller window areas
[@lyonsWindowPerformanceHuman1999]. Many of these scenarios would be
acceptable under the covered-ankle model but flagged as problematic
under the current single-threshold approach.

For practitioners, this suggests a two-criteria framework: apply the
existing model (or the uncovered-ankle model from this study) when
evaluating spaces where occupants are likely to have exposed ankles, and
apply the covered-ankle model for winter heating scenarios where long
trousers, socks, and closed-toe shoes are expected. Within the
near-neutral thermal range where ASHRAE Standard 55 applies, PMV
provides the best available proxy for whole-body thermal sensation
during design evaluation. These implications relate specifically to the
convective ankle draft component; radiant asymmetry from cold glazing,
governed by a separate criterion in ASHRAE Standard 55 (Section 5.3),
should be evaluated independently and may independently necessitate
perimeter heating near large or poorly insulated windows.

The practical benefits are significant. Perimeter heating systems,
increased façade insulation and other techniques used to reduce the
downdraft represent non-trivial costs and carbon emissions in both
installation and operation. If covered and uncovered conditions are
evaluated separately, designers may be able to downsize or remove
solutions used to manage the problem. This approach could yield
meaningful reductions in first costs, energy consumption, and associated
carbon emissions while maintaining occupant comfort.

It should also be noted that the proposed models are not limited to
window downdraft scenarios: since the inputs are ankle-level air speed,
air temperature, and whole-body thermal sensation — all directly
measurable quantities — the models are equally applicable to other
low-velocity air distribution strategies producing similar ankle-level
conditions, including underfloor air distribution and displacement
ventilation systems.

## Limitations

Several limitations should be considered when interpreting these
findings.

First, this study was conducted in a controlled laboratory environment
with standardized clothing, which differs from real settings where
occupants exhibit considerable variability in attire even during winter
months. While we targeted 0.75 clo to represent median winter clothing,
actual insulation levels vary with individual preferences, building
temperatures, and cultural norms. The protective effect of ankle
coverage may differ with clothing types not tested here, such as
ankle-length skirts, cropped trousers, or low-cut shoes.

Second, the participant sample consisted primarily of
university-affiliated individuals within a limited age range. Thermal
sensitivity varies with age, and older occupants may respond differently
to ankle draft. The generalizability of these findings to broader
occupant populations warrants further investigation.

Third, to enable a comparison with Liu’s study, the experimental
protocol used the same 20 min exposure duration, whereas real occupants
may experience sustained exposure over several hours. Although the
extrapolation analysis suggested that the effect of longer exposure on
the main conclusions may be limited, these estimates remain
assumption-based. Air movement acceptability beyond the 20 min exposure
period could not be directly predicted. Therefore, the percentage of
occupants dissatisfied with ankle draft under longer exposure durations
may be higher than that estimated by the current model. Similarly,
participants remained seated throughout the experiment; standing or
walking occupants may experience different ankle-level thermal
conditions.

Fourth, the experimental conditions covered supply air temperatures of
15–19°C, leading to ankle-level air temperatures from 17-20°C, and air
speeds of 0.1–0.7 m/s. While these ranges were selected to represent
typical window downdraft conditions, more extreme scenarios (colder
supply/ankle-level temperatures or higher velocities) were not tested.
The model predictions should be applied with caution outside the tested
parameter space.

Fifth, airflow was delivered from behind the participants to simulate a
window downdraft approaching their lower legs from the façade. This
scenario, identified by @toftumEffectAirflowDirection1997 as one of the
most unfavorable situations, was used to study the impact of airflow on
the participants. Other airflow directions (lateral or frontal) may
produce different perceptual responses, particularly given potential
differences in skin exposure and attention to the stimulus source.

Finally, the experimental setup simulated the convective component of
window downdraft but did not include radiant asymmetry from a cold
window surface. In real buildings, occupants near poorly insulated
glazing experience both convective cooling from downdraft and radiant
heat loss to the cold surface. The combined effect of these two
mechanisms on ankle-level comfort was beyond the scope of this study.

# Conclusions

This study evaluated thermal comfort responses to ankle-level draft
under typical winter clothing conditions. Through controlled laboratory
experiments with participants wearing approximately 0.75 clo ensembles
that covered the ankle, we characterized local and whole-body thermal
perception across a range of ankle level air temperatures (17-20°C) and
air speeds (0.1–0.7m/s) simulating window downdraft.

The principal findings indicate that the existing ankle draft model
systematically overpredicts dissatisfaction under winter clothing
conditions. While prior studies with exposed ankles reported
dissatisfaction rates between 23% and 57% under similar air speeds, our
observations with covered ankles ranged from only 6% to 20%. This
suggests that ankle coverage substantially reduces sensitivity to local
draft, a conclusion supported by the covered-ankle model developed in
this study; the model demonstrates both a lower baseline risk and a
reduced response to air speed compared to uncovered-ankle models, with
an intercept difference of
`r sprintf("%.1f", abs(glmm_delta_intercept))` units on the logit scale.

Consistent with prior model structures, we found that whole-body thermal
sensation and air speed at the ankle are the primary predictors of
dissatisfaction, whereas ankle air temperature, room temperature, and
turbulence intensity were not significant within the tested ranges.

This reduced sensitivity under winter clothing appears to stem from a
combination of physiological mechanisms and potentially psychological
factors. Physiologically, increased thermal resistance from clothing
coverage likely attenuated the effect of local cooling, with lower-leg
skin temperature decreasing by 0.8°C to 1.7°C across all experimental
conditions. Psychologically, perceived protection from clothing may also
have contributed to the reduced sensitivity.

These findings support a differentiated approach to ankle draft
assessment in building standards. For winter heating scenarios where
occupants typically wear clothing that covers the ankle (≥0.75 clo), the
covered-ankle model provides more accurate predictions than the current
ASHRAE Standard 55 criterion. Adopting separate criteria for covered and
uncovered conditions could enable significant reductions in perimeter
heating requirements without compromising occupant comfort, yielding
benefits in installation costs, energy consumption, and carbon
emissions.

# CRediT authorship contribution statement {.unnumbered}

**Tobias Kramer**: Conceptualization, Data curation, Formal analysis,
Methodology, Investigation, Project administration, Software, Writing -
original draft, Writing - review & editing, Visualization. **Junmeng
Lyu**: Data curation, Formal analysis, Methodology, Investigation,
Software, Writing - original draft, Writing - review & editing,
Visualization. **Stefano Schiavon**: Conceptualization, Funding
acquisition, Methodology, Project administration, Supervision, Writing -
review & editing, Visualization.

# Declaration of competing interest {.unnumbered}

All authors declare no competing interest. The Center for the Built
Environment (CBE) at the University of California, Berkeley, with which
the authors are affiliated, is advised and funded in part by
approximately 50 partners that represent a diversity of organizations
from the building industry, including manufacturers, building owners,
facility managers, contractors, architects, engineers, government
agencies, and utilities.

# Data availability {.unnumbered}

The datasets generated and analyzed during the current study as well as
the full source code of the data analysis workflow are available in the
following GitHub repository:
[https://github.com/CenterForTheBuiltEnvironment/ankle-draft](https://github.com/CenterForTheBuiltEnvironment/ankle-draft-II){.uri}

# Declaration of Generative AI and AI-assisted technologies in the writing process {.unnumbered}

During the preparation of this work, the authors utilized the chatbot
Claude to enhance the language and readability of the content.
Subsequently, they reviewed and edited the material as required, while
maintaining full accountability for the publication’s content.

# Acknowledgements {.unnumbered}

This research was supported by Schüco and the Center for the Built
Environment (CBE) at the University of California, Berkeley. The authors
express their gratitude to Kit Elsworth for his valuable input and
leadership throughout the project.

# Appendix A: Response distributions {#sec-appendix-a .unnumbered}

The following figures show the full response distributions for thermal
sensation, thermal preference, and air movement acceptability and
preference. These complement the scatter plots with mixed-effects
regression lines presented in @sec-results in the main text.
@tbl-supp-pairwise-perception summarizes the condition comparisons that
showed statistically significant differences.

![Whole-body thermal sensation (a) and thermal preference (b) votes by
ankle-level air temperature and air speed. Participants predominantly
reported neutral to cool sensations, with lower ankle-level air
temperatures increasing the proportion preferring warmer
conditions.](figs/supp_thermal_perception_overall.png){#fig-supp-thermal-perception-overall
width="75%"}

![Ankle-level thermal sensation (a) and thermal preference (b) votes by
ankle-level air temperature and air speed. Sensations were generally
cooler at the ankles than whole-body ratings, with lower ankle-level air
temperatures and higher air speeds increasing cool sensations and the
preference for warmer
conditions.](figs/supp_thermal_perception_ankle.png){#fig-supp-thermal-perception-ankle
width="75%"}

![Ankle-level air movement acceptability (a) and preference (b) votes by
ankle-level air temperature and air speed. Most participants found ankle
air movement acceptable, though higher air speeds increased the
preference for less air
movement.](figs/supp_air_movement_acc_pref.png){#fig-supp-air-movement-acc-pref
width="75%"}

```{r}
#| label: tbl-supp-pairwise-perception
#| tbl-cap: "Significant pairwise comparisons for perception responses shown in Appendix A."
#| echo: false

supp_pairwise_perception_table <- dplyr::bind_rows(
  thermal_sensation_overall_stats %>%
    dplyr::mutate(Perception = "Thermal sensation, whole body"),
  thermal_preference_overall_stats %>%
    dplyr::mutate(Perception = "Thermal preference, whole body"),
  thermal_sensation_ankles_stats %>%
    dplyr::mutate(Perception = "Thermal sensation, ankles"),
  thermal_preference_ankles_stats %>%
    dplyr::mutate(Perception = "Thermal preference, ankles"),
  air_movement_acceptability_stats %>%
    dplyr::mutate(Perception = "Air movement acceptability, ankles"),
  air_movement_preference_stats %>%
    dplyr::mutate(Perception = "Air movement preference, ankles")
) %>%
  dplyr::filter(!is.na(p.adj), p.adj < 0.05) %>%
  dplyr::transmute(
    Perception,
    Comparison,
    Condition,
    `Pairwise comparison`,
    n,
    `Mean group 1` = mean1,
    `Mean group 2` = mean2,
    `Adjusted p` = sprintf("%.3f", p.adj),
    `Effect size r` = sprintf("%.3f", r)
  )

knitr::kable(
  supp_pairwise_perception_table,
  align = c("l", "l", "l", "l", "r", "r", "r", "r", "r")
)
```

# Appendix B: Monte Carlo approximation of marginal predicted probabilities {#sec-appendix-b .unnumbered}

In the fitted generalized linear mixed model, the inverse-logit
transformation is nonlinear. Therefore, directly applying the
inverse-logit transformation to the fixed-effects linear predictor gives
a conditional probability, corresponding to the predicted probability
when the subject-level random intercept is set to zero. This quantity
does not generally equal the population-averaged probability because the
average of inverse-logit-transformed values over the random-effect
distribution is not equal to the inverse-logit transformation of the
average linear predictor.

To obtain population-averaged predictions, we estimated marginal
predicted probabilities by integrating over the random-intercept
distribution. Specifically, fixed-effect estimates and the variance of
the subject-level random intercept were extracted from the fitted model.
The random intercept was assumed to follow a normal distribution with
mean zero and the estimated variance from the fitted model. For each
combination of predictor values, random-intercept values were sampled
from this distribution, and the predicted probability was calculated
after applying the inverse-logit transformation. The marginal predicted
probability was then approximated by averaging these probabilities
across Monte Carlo simulation (K = 10,000, random seed = 1).

Because these Monte Carlo-based marginal probabilities generally do not
have a simple closed-form expression, we fitted a logistic-form
approximation to the simulated marginal probabilities, with a linear
predictor on the logit scale. This approximation was used to provide a
practical population-averaged risk equation. The reported approximation
$R^2$ values should not be interpreted as the explanatory power of the
original mixed-effects model. Instead, they quantify the agreement
between the closed-form approximation and the Monte Carlo-based marginal
predictions.

# Appendix C: Effect of draft dissatisfaction definition on model {#sec-appendix-c .unnumbered}

```{r}
#| label: appendix-c-definition-values
#| echo: false
#| message: false

appendix_c_rates <- new_model %>%
  dplyr::filter(
    (source == "toby" & clothing_type == "long") |
      (source == "liu" & clothing_type == "short")
  ) %>%
  dplyr::mutate(group = dplyr::if_else(source == "toby", "toby_covered", "liu_uncovered")) %>%
  dplyr::group_by(group) %>%
  dplyr::summarise(
    acc = mean(dissatisfied_with_draft_ankles, na.rm = TRUE),
    pref = mean(dissatisfied_with_draft_ankles_amp, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  tidyr::pivot_wider(names_from = group, values_from = c(acc, pref))

format_pct_1 <- function(x) {
  scales::percent(x, accuracy = 0.1)
}

v20 <- function(model_formula_tbl, model_name, ts, p = 0.20) {
  row <- model_formula_tbl %>% dplyr::filter(Model == model_name)
  (qlogis(p) - row$Intercept - row$TS_coef * ts) / row$V_coef
}

format_v20 <- function(v) {
  dplyr::case_when(
    v < 0 ~ "even at 0 m/s",
    v > 1 ~ "above 1.00 m/s",
    TRUE ~ paste0(sprintf("%.2f", v), " m/s")
  )
}

appendix_c_thresholds <- tidyr::expand_grid(
  definition = c("acc", "pref"),
  Model = c("exposed", "unexposed"),
  TS = c(-0.5, 0.5)
) %>%
  dplyr::mutate(
    V20 = dplyr::case_when(
      definition == "acc" ~ purrr::map2_dbl(
        Model, TS, \(model_name, ts) v20(model_formula, model_name, ts)
      ),
      definition == "pref" ~ purrr::map2_dbl(
        Model, TS, \(model_name, ts) v20(model_formula_amp, model_name, ts)
      )
    ),
    V20_text = format_v20(V20)
  )

get_v20_text <- function(definition, model_name, ts) {
  appendix_c_thresholds %>%
    dplyr::filter(definition == !!definition, Model == !!model_name, TS == !!ts) %>%
    dplyr::pull(V20_text)
}

appendix_c_pref_covered_range <- appendix_c_thresholds %>%
  dplyr::filter(definition == "pref", Model == "unexposed") %>%
  dplyr::summarise(range_text = paste0(sprintf("%.2f", min(V20)), "-", sprintf("%.2f", max(V20)), " m/s")) %>%
  dplyr::pull()
```

Draft is defined as unwanted local cooling of the body caused by air
movement [@ashraeANSIASHRAEStandard2023]. To enable comparison with the
existing model, this study adopted the same definition of draft
dissatisfaction as @liuPredictedPercentageDissatisfied2017, where an
occupant is considered dissatisfied with draft when the local thermal
sensation is below 0 and the local air movement is unacceptable.

This definition raises a methodological issue: whether draft
dissatisfaction should be characterized using air movement acceptability
or air movement preference. Compared with acceptability, preference is
stricter. A preference for less air movement does not necessarily mean
that the current condition is unacceptable; rather, it may indicate a
demand for a higher level of comfort. Under the ankle-covered condition
in this study, the incidence of draft dissatisfaction was approximately
`r format_pct_1(appendix_c_rates$acc_toby_covered)` when defined as
“local thermal sensation \< 0 and air movement unacceptable.” When
defined as “local thermal sensation \< 0 and preference for less air
movement,” the incidence increased to approximately
`r format_pct_1(appendix_c_rates$pref_toby_covered)`. A similar pattern
was observed in the ankle-uncovered subset of Liu's dataset, where the
observed dissatisfaction rate increased from
`r format_pct_1(appendix_c_rates$acc_liu_uncovered)` to
`r format_pct_1(appendix_c_rates$pref_liu_uncovered)` when the
preference-based definition was used.

To further examine the influence of these two definitions on model
outcomes, we fitted the draft dissatisfaction model using the same model
structure, but redefining draft dissatisfaction as “local thermal
sensation \< 0 and air movement preference = less.” The results are
shown in @fig-model-amp.

![Prediction surfaces of draft dissatisfaction based on the
preference-based definition. Draft dissatisfaction was defined as local
thermal sensation below 0 and a preference for less air movement. (a)
ankle-uncovered conditions, and (b) ankle-covered
conditions.](figs/Model_amp.png){#fig-model-amp width="100%"}

The results show clear differences between the limits derived from the
two definitions, especially around the 20% dissatisfaction threshold.
For whole-body thermal sensation between −0.5 and +0.5, the
acceptability-based model gives a threshold air speed range of
`r get_v20_text("acc", "exposed", -0.5)` to
`r get_v20_text("acc", "exposed", 0.5)` under the ankle-uncovered
condition, and `r get_v20_text("acc", "unexposed", -0.5)` to
`r get_v20_text("acc", "unexposed", 0.5)` under the ankle-covered
condition. In contrast, the preference-based model shifts the predicted
dissatisfaction percentage higher and narrows the allowable air speed
range. Under the ankle-uncovered condition, the 20% threshold is
exceeded at TS = −0.5 `r get_v20_text("pref", "exposed", -0.5)`, and the
threshold air speed is only `r get_v20_text("pref", "exposed", 0.5)` at
TS = +0.5. Under the ankle-covered condition, the threshold range is
approximately `r appendix_c_pref_covered_range`, which is still lower
than that obtained using the acceptability-based definition.

Overall, the two definitions do not alter the main conclusion of the
model, which is ankle coverage reduces the risk of draft
dissatisfaction. However, the acceptability-based definition is more
appropriate for defining standard limits. In contrast, the
preference-based definition provides more conservative air speed limits
and reflects the need for further improvement in air movement conditions
under a higher comfort requirement.

# References {.unnumbered}
