---
title: "Thermal comfort responses to ankle-level draft under winter clothing conditions"
authors:
  - name: Tobias Kramer
    affiliations:
      - ref: cbe
    corresponding: true
    email: t.kramer@berkeley.edu
    orcid: 0000-0000-0000-0000
  - name: Junmeng Lyu
    affiliations:
      - ref: cbe
      - ref: sjtu
    orcid: 0000-0000-0000-0000
  - name: Stefano Schiavon
    affiliations:
      - ref: cbe
      - ref: arch
      - ref: eng
    orcid: 0000-0003-1285-5682 

affiliations:
  - id: cbe
    name: Center for the Built Environment, University of California, Berkeley, USA
  - id: arch
    name: Department of Architecture, College of Environmental Design, University of California, Berkeley, USA
  - id: eng
    name: Department of Civil and Environmental Engineering, College of Engineering, University of California, Berkeley, USA
  - id: sjtu
    name: Department of Architecture, School of Design, Shanghai Jiao Tong University, Shanghai, China

filters:
  - authors-block

date: "`r Sys.Date()`"
format:
  docx:
    link-citations: true
    reference-doc: template.docx
    toc: false
    number-sections: true
  html:
    toc: true
    number-sections: true
    theme: cosmo
    comments:
      hypothesis: true
bibliography: refs.bib
csl: apa.csl
execute:
  echo: false
  warning: false
  error: false
---

```{r load-main-analysis}
#| message: false
#| warning: false
#| paged-print: false
source(here::here("src", "R", "x_analysis.R"))
```

# Abstract {.unnumbered}

<!-- Draft after results are finalized -->

# Introduction

Cold downdraft near windows is a common source of local thermal
discomfort in buildings during winter. When outdoor temperatures drop,
the interior surface of windows, particularly large or poorly insulated
glazing, becomes significantly cooler than the surrounding air. This
temperature differential causes nearby air to cool, increase in density,
and descend, creating a localized airflow pattern that can cause
discomfort at the lower extremities of seated occupants
[@heiselbergDraftComfortReview1994
<!--# Replace reference with BetterBibTex -->]. Building designers
typically address this issue by installing perimeter heating systems,
such as radiators or in-floor heating, to counteract the downdraft
effect and to prevent thermal discomfort in the ankle region.

The current approach to evaluating ankle draft risk relies on a
predictive model presented in @liuPredictedPercentageDissatisfied2017,
which estimates the predicted percentage dissatisfied (PPD) with ankle
draft as a function of air speed and temperature at ankle level, as well
as whole-body thermal sensation. This model, now incorporated into
ASHRAE Standard 55 [@ashraeANSIASHRAEStandard2023], was derived from
laboratory experiments conducted under cooling conditions typical of
underfloor air distribution (UFAD) and displacement ventilation systems.
Importantly, participants in these foundational studies, including an
earlier investigation by @schiavonSensationDraftUncovered2016, wore
summer clothing with exposed ankles, resulting in clothing insulation
levels of approximately 0.5 clo.

However, building occupants in winter typically wear substantially
different attire. Analysis of the ASHRAE Global Thermal Comfort Database
II [@foldvarylicinaDevelopmentASHRAEGlobal2018] indicates that median
clothing insulation in office environments during heating seasons is
approximately 0.75 clo, with occupants commonly wearing long trousers,
closed-toe shoes and socks that cover the ankle region. This discrepancy
raises an important question: does the existing ankle draft model,
calibrated for summer conditions with exposed skin, accurately predict
discomfort when ankles are covered by clothing?

If the current model overestimates draft risk under winter clothing
conditions, the practical implications are significant. Designers may be
specifying perimeter heating systems that are larger than necessary, or
installing them in situations where they could be avoided entirely. This
has consequences for both first costs and operational energy
consumption, as perimeter heating systems contribute to building energy
use and associated carbon emissions. Conversely, if the model
underestimates risk, occupants may experience discomfort that
compromises satisfaction and productivity.

The objective of this study is to evaluate thermal comfort responses to
ankle-level draft under typical winter clothing conditions through
controlled laboratory experiments. Specifically, we aim to: (1)
characterize whole-body and local thermal perception across a range of
supply air temperatures and air speeds representative of window
downdraft conditions; (2) compare observed dissatisfaction rates with
predictions from the existing ankle draft model; and (3) assess whether
the current ASHRAE 55 criteria require adjustment for winter scenarios.
The findings are intended to inform guidance on perimeter heating
requirements and support more energy-efficient building design without
compromising occupant comfort.

# Methods {#sec-methods}

## Experimental design

We conducted a controlled laboratory experiment using a within-subjects
repeated-measures design. Each participant was exposed to nine
experimental conditions in a 3×3 factorial arrangement, varying supply
air temperature (SAT) at three levels (15°C, 17°C, 19°C) and air speed
at ankle level at three levels (low, medium, high; ranging between
0.1–0.7 m/s). The SAT and air speeds were selected to represent the
range of conditions associated with window downdraft under typical
winter conditions, informed by computational fluid dynamics simulations
conducted as part of a broader research collaboration.

PLACEHOLDER EXPERIMENTAL PROTOCOL GRAPHIC

Participants attended four sessions: an introductory session for
explaining the experiment and obtaining consent, followed by three
experimental sessions of approximately two hours each. Each experimental
session corresponded to one SAT level, with participants experiencing
all three air speed conditions within that session in a counterbalanced
order. We scheduled sessions on separate days to minimize carryover
effects and subjects were allowed to attend the three sessions in their
preferred order.

## Climate chamber and apparatus

We conducted the experiment in a climate-controlled chamber at the
Center for the Built Environment, UC Berkeley. The chamber floor area is
approximately 25 m². We set up three workstations to allow simultaneous
testing of multiple participants and three seats for reacclimatization
during breaks in the back of the room. Ambient room conditions were
maintained at approximately 22°C operative temperature, 50% relative
humidity, and background air velocity below 0.1 m/s, targeting a
whole-body thermal sensation of approximately −0.5 (slightly cool) on
the ASHRAE seven-point scale as calculated using the CBE Comfort Tool
[@tartariniCBEThermalComfort2020].

PLACEHOLDER CHAMBER SETUP GRAPHIC

We used custom displacement diffusers to deliver conditioned air to the
ankle region of participants seated at the three workstations. The
diffuser was positioned 0.7 m behind each workstation, directing airflow
toward the participants' lower legs from behind and simulating the
approach direction of window downdraft. The diffuser was connected to an
independent air handling unit capable of delivering supply air at the
target temperatures and flow rates. The three supply air temperatures
(15°C, 17°C, 19°C) represent temperatures at the VAV outlet and we
controlled air flow rates by damper adjustment. Turbulence intensity at
the measurement location was approximately 30% (see @tbl-env),
consistent with values reported in prior ankle draft studies
[@liuPredictedPercentageDissatisfied2017]. The three seats in the back
of the room were not affected by the diffusers targeting the three
workstations only.

## Monitoring

### Environmental

We measured head-level air temperature (1.1 m) at each workstation using
calibrated Atmocube monitors (Atmotech Inc., United States; air
temperature: -40 to +125°C, ± 0.5°C, relative humidity: 0 to 100%RH, ± 2
%RH, CO~2~: 0 to 5000 ppm, ± 50 ppm + 2.5% of reading at standard room
levels) positioned within 0.5 m of participants. To measure air velocity
and air temperature at ankle level (0.1 m) we used the AirDistSys 5000
omnidirectional hot-wire anemometers (Sensor Electronic, Poland, air
velocity: 0.05-5 m/s, accuracy: ±0.02 m/s ±2% of reading; temperature:
-10 to +50°C, accuracy: ±0.2°C, sampling interval: 1 s). Room-level
measurements of relative humidity and CO~2~ concentration were recorded
throughout each session.

### Physiological

In addition, we recorded continuously participants skin temperature on
the lateral lower leg (mid-calf) and feet using iButton sensors
(Thermochron, Maxim Integrated, USA, Type DS1923; measurement range:
20°C to 85°C; accuracy: ±0.5°C; resolution: 0.0625°C) attached with
medical-grade hypoallergenic tape. We used the skin temperature data to
assess local cooling responses and as physiological correlates of
subjective thermal sensation.

### Subjective feedback

To assess the perception of the thermal environment we used validated
scales consistent with prior ankle draft research and ASHRAE Standard 55
requirements. Constructs and their scales used in the questionnaires are
summarized in @tbl-scales below. We administered the questions
electronically via a custom survey tool written in Java. Participants
were not required to respond to any individual question, though
completion rates were monitored.

```{r}
#| label: tbl-scales
#| tbl-cap: "Survey constructs and measurement scales"
#| tbl-colwidths: [15,25,15,45]
#| echo: false

readr::read_csv("tables/survey_scales.csv") |>
  knitr::kable()
```

## Participants

We recruited
`r demographic_d %>% dplyr::filter(sex == "All") %>% dplyr::pull(n)`
participants from the university community and local area through the
Experimental Social Science Laboratory (Xlab) platform and email
outreach to individuals who had participated in previous studies and
consented to future contact. Eligible participants were between 21 and
55 years of age, fluent in English, and free from chronic
cardiorespiratory conditions or recent major surgery. Pregnant
individuals were excluded. Demographic details of the cohort are
summarized in @tbl-demographics.

## Experimental protocol

Upon arrival for each experimental session, we checked participants for
elevated temperature and fitted them with skin temperature sensors.
Participants then entered the climate chamber and completed a 20-minute
adaptation period in the acclimatization zone under baseline conditions.

To simulate typical winter office attire, we instructed participants to
wear a long-sleeve top, long trousers (ending just above the ankle),
thin socks, and closed-toe shoes. This ensemble targets a clothing
insulation of approximately 0.75 clo, consistent with the median value
observed in office buildings during heating seasons
[@foldvarylicinaDevelopmentASHRAEGlobal2018]. Participants were
permitted to make minor clothing adjustments during sessions (e.g.,
rolling sleeves), and any changes were recorded.

Following adaptation, participants were exposed to three consecutive
20-minute conditions at different air speed levels, with 20-minute rest
periods between conditions (see PROTOCOL GRAPHIC REFERENCE). The
sequence of air speed conditions was counterbalanced across participants
and sessions. During each exposure period, participants completed
standardized questionnaires at the end of each interval.

## Statistical analysis

### Variable processing

Thermal sensation votes were collected on a continuous scale. For
descriptive analyses, we mapped these values to the corresponding
categorical sensation levels, whereas the original continuous values
were retained for statistical modeling. We dichotomized ankle airflow
acceptability as acceptable or unacceptable for subsequent analyses. The
original recorded votes for thermal acceptability, thermal preference,
ankle thermal preference, and ankle airflow preference were retained
without further processing. Consistent with the definition used by
@liuPredictedPercentageDissatisfied2017, we defined ankle draft
dissatisfaction as an ankle thermal sensation vote below 0 (neutral)
together with an ankle airflow acceptability vote below 0.

### Within-subject comparisons

We used a within-subject repeated-measures design and conducted
comparisons between experimental conditions within participants. We
analyzed continuous variables using paired-samples t-tests and reported
paired Cohen's d as the effect size. We analyzed ordinal variables using
the paired Wilcoxon signed-rank test, reporting the test statistic based
on the normal approximation and the effect size r. For multiple pairwise
comparisons within each family of tests, we adjusted p-values using the
Benjamini-Hochberg method \[\@benjaminiHighberg1995\].

### Evaluation of the existing ankle draft model

Applying it to the experimental data, we calculated the predicted
percentage of dissatisfaction with ankle draft at each observation
point. Predictions were then averaged by experimental condition and
compared with the observed dissatisfaction percentages. We used
calibration curves to quantitatively assess the model's predictive
performance in this study's dataset \[\@vancalster2016\] and report the
calibration intercept and slope, AUC, and Brier score. All analyses were
implemented using the R package CalibrationCurves
\[\@deCockCalibrationCurves2023\].

### Development of the new ankle draft model

We modeled ankle draft dissatisfaction using a generalized linear mixed
model with a binomial distribution and logit link, as the outcome was
binary and repeated observations were correlated within subjects. We
included the subject as a random intercept to account for
between-subject heterogeneity and within-subject dependence arising from
repeated measurements. To enable an integrated analysis of the effects
of different clothing conditions on ankle draft dissatisfaction, we also
included data from @liuPredictedPercentageDissatisfied2017 in the
modeling dataset. Continuous predictors were mean-centered using the
overall means of the combined dataset before model fitting.

In this generalized linear mixed model, the inverse-logit transformation
is nonlinear. As a result, direct back-transformation of the
fixed-effects linear predictor yields a conditional probability,
representing an individual-level prediction when the random effect is
set to zero, rather than a population-averaged probability. To
characterize the average risk of dissatisfaction at the population
level, we therefore use marginal probabilities instead of conditional
probabilities [@neuhausKalbfleischHauck1991]. Specifically, we extracted
fixed-effect estimates and the variance of the random intercept from the
fitted model, assuming that the random intercept followed a normal
distribution with mean zero. We then approximated integration over the
random effects using Monte Carlo simulation (K = 10,000, random seed =
1), and calculated the marginal predicted probability for each
combination of predictor values.

Because these marginal probabilities generally cannot be written as a
simple closed-form equation, we used the Monte Carlo-simulated marginal
probabilities to fit an approximate linear equation for logit(P).
Applying the inverse-logit transformation to the fitted equation yielded
a closed-form prediction equation. We evaluated the accuracy of this
approximation using coefficient of determination (R^2^).

## Software

We performed all data processing, statistical analyses, and model
fitting in R (version 4.4.1). The main packages we used were tidyverse
(version 2.0.0), lme4 (version 2.0.1), rstatix (version 0.7.3), and
CalibrationCurves (version 3.0.0).

# Results

## Participants and environmental conditions

```{r}
#| label: tbl-demographics
#| tbl-cap: "Summary of demographic information (mean ± SD) on study participants."

knitr::kable(demographic_d,
             col.names = c('Sex', 'No. Subjects', 'Age [a]', 'Height [m]', 'Weight [kg]', 'Body Mass Index (BMI)')
)

```

@tbl-demographics summarizes the demographic information of the study
subjects. A total of
`r demographic_d %>% dplyr::filter(sex == "All") %>% dplyr::pull(n)`
subjects participated in the study
(`r demographic_d %>% dplyr::filter(sex == "Female") %>% dplyr::pull(n)`
female,
`r demographic_d %>% dplyr::filter(sex == "Male") %>% dplyr::pull(n)`
male,
`r demographic_d %>% dplyr::filter(sex == "Third Gender / Other") %>% dplyr::pull(n)`
other), with a mean age of
`r demographic_d %>% dplyr::filter(sex == "All") %>% dplyr::pull(age)`
years.

```{=html}
<!--
- Final sample size, completion rate, demographics (age, sex distribution)
- Table: Achieved environmental conditions by SAT level and air speed condition
  (mean ± SD for air temp, supply temp, air velocity at ankle, RH)
- Brief statement confirming conditions were achieved as designed
-->
```

```{r}
#| label: tbl-env
#| tbl-cap: "Summary of environmental measurements."

knitr::kable(env_airflow_t,
             col.names = c('Measurement', '15C', '17C', '19C')
)
```

@tbl-env provides an overview of the measured environmental conditions.
Across all three sessions, head-level air temperature remained stable at
`r env_airflow_t %>% dplyr::filter(measurement == "Air temperature, head level (°C)") %>% dplyr::pull("15C")`
°C,
`r env_airflow_t %>% dplyr::filter(measurement == "Air temperature, head level (°C)") %>% dplyr::pull("17C")`
°C, and
`r env_airflow_t %>% dplyr::filter(measurement == "Air temperature, head level (°C)") %>% dplyr::pull("19C")`
°C, and relative humidity was comparable across sessions, confirming
stable background thermal conditions.

As the primary experimental manipulation, we set supply air temperatures
to 15°C, 17°C, and 19°C to simulate the effect of window downdraft.
Ankle-level temperatures were slightly higher than the supply
temperature due to mixing with the warmer ambient air, but remained
stable across all sessions. Within each session, we exposed subjects to
three airspeed levels at separate workstations. We measured consistent
air speeds across all three workstations and sessions, confirming
reliable reproduction of the low, medium, and high airspeed conditions.
Turbulence intensities remained low throughout, indicating stable and
controlled airflow.

## Whole body thermal perception

```{r}
#| label: whole-body-helpers
#| echo: false

# Helper functions for inline text
get_sensation_pct <- function(sat, ws, response) {
  thermal_sensation_overall_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

get_preference_pct <- function(sat, ws, response) {
  thermal_preference_overall_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

# Aggregated summaries for text
sensation_neutral_cool <- thermal_sensation_overall_summary %>%
  dplyr::filter(response_value %in% c("Neutral", "Slightly cool", "Cool")) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(pct = sum(pct), .groups = "drop")

preference_no_change <- thermal_preference_overall_summary %>%
  dplyr::filter(response_value == "No change")

preference_warmer <- thermal_preference_overall_summary %>%
  dplyr::filter(response_value == "Warmer")
```

@fig-thermal-perception-overall presents whole-body thermal sensation
and preference responses across all experimental conditions. Thermal
sensation votes were predominantly in the neutral to slightly cool
range, with
`r sensation_neutral_cool %>% dplyr::summarise(range = paste0(scales::percent(min(pct)), "–", scales::percent(max(pct)))) %>% dplyr::pull(range)`
of participants reporting sensations between "Cool" and "Neutral" across
conditions. At the lowest supply air temperature (15°C),
`r get_sensation_pct("15C", "high", "Slightly cool")` of participants
reported feeling slightly cool at the highest air speed, compared to
`r get_sensation_pct("15C", "low", "Slightly cool")` at the lowest air
speed. This pattern was less pronounced at warmer supply temperatures.

Thermal preference responses indicated that participants were generally
satisfied with the thermal conditions. Across all conditions,
`r preference_no_change %>% dplyr::summarise(range = paste0(scales::percent(min(pct)), "–", scales::percent(max(pct)))) %>% dplyr::pull(range)`
of participants preferred "No change" to the thermal environment. The
proportion preferring warmer conditions ranged from
`r preference_warmer %>% dplyr::summarise(range = paste0(scales::percent(min(pct)), " to ", scales::percent(max(pct)))) %>% dplyr::pull(range)`,
with the highest values observed during the 15°C supply air temperature
sessions.

![Thermal sensation (a) and thermal preference (b) for whole
body.](figs/thermal_perception_overall.png){#fig-thermal-perception-overall
width="75%"}

```{r}
#| label: acceptability-helpers
#| echo: false

# Thermal acceptability summary
thermal_acceptability_summary <- thermal_acceptability_d %>%
  dplyr::mutate(acceptable = response_value > 0) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(
    n = dplyr::n(),
    pct_acceptable = mean(acceptable),
    mean_value = mean(response_value),
    .groups = "drop"
  )

acceptability_range <- thermal_acceptability_summary %>%
  dplyr::summarise(
    min = scales::percent(min(pct_acceptable), accuracy = 1),
    max = scales::percent(max(pct_acceptable), accuracy = 1)
  )

acceptability_by_sat <- thermal_acceptability_summary %>%
  dplyr::group_by(session_sat) %>%
  dplyr::summarise(
    min = scales::percent(min(pct_acceptable), accuracy = 1),
    max = scales::percent(max(pct_acceptable), accuracy = 1),
    .groups = "drop"
  )
```

@fig-thermal-acceptability-overall shows thermal acceptability responses
for whole-body conditions. Overall, acceptability was high across all
conditions, ranging from `r acceptability_range$min` to
`r acceptability_range$max`. Acceptability rates were highest during the
19°C supply air temperature sessions
(`r acceptability_by_sat %>% dplyr::filter(session_sat == "19C") %>% dplyr::mutate(range = paste0(min, "–", max)) %>% dplyr::pull(range)`)
and lowest at 15°C
(`r acceptability_by_sat %>% dplyr::filter(session_sat == "15C") %>% dplyr::mutate(range = paste0(min, "–", max)) %>% dplyr::pull(range)`).
Within each supply air temperature condition, air speed had minimal
effect on whole-body thermal acceptability.

![Thermal acceptability for whole
body.](figs/thermal_acceptability_overall.png){#fig-thermal-acceptability-overall
width="100%"}

## Ankle level observations

### Thermal perception

```{r}
#| label: ankle-helpers
#| echo: false

# Helper functions for ankle-level inline text
get_ankle_sensation_pct <- function(sat, ws, response) {
  thermal_sensation_ankle_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

get_ankle_preference_pct <- function(sat, ws, response) {
  thermal_preference_ankle_summary %>%
    dplyr::filter(session_sat == sat, workstation == ws, response_value == response) %>%
    dplyr::pull(pct_label)
}

# Aggregated ankle sensation on cool side (cold + cool + slightly cool)
ankle_sensation_cool_side <- thermal_sensation_ankle_summary %>%
  dplyr::filter(response_value %in% c("Cold", "Cool", "Slightly cool")) %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(pct = sum(pct), .groups = "drop")

# Ankle preference for warmer
ankle_preference_warmer <- thermal_preference_ankle_summary %>%
  dplyr::filter(response_value == "Warmer")
```

@fig-thermal-perception-ankle presents thermal sensation and preference
responses at the ankle level. Compared to whole-body responses,
participants reported cooler sensations at the ankles. The proportion
reporting sensations on the cool side of neutral (cold, cool, or
slightly cool) ranged from
`r ankle_sensation_cool_side %>% dplyr::summarise(range = paste0(scales::percent(min(pct)), " to ", scales::percent(max(pct)))) %>% dplyr::pull(range)`
across conditions. At the coldest supply air temperature (15°C) with
high air speed,
`r ankle_sensation_cool_side %>% dplyr::filter(session_sat == "15C", workstation == "high") %>% dplyr::mutate(pct = scales::percent(pct)) %>% dplyr::pull(pct)`
of participants reported cool sensations at the ankles, compared to
`r ankle_sensation_cool_side %>% dplyr::filter(session_sat == "19C", workstation == "low") %>% dplyr::mutate(pct = scales::percent(pct)) %>% dplyr::pull(pct)`
at 19°C with low air speed.

Thermal preference at the ankle level reflected this cooling effect. The
proportion of participants preferring warmer ankles ranged from
`r ankle_preference_warmer %>% dplyr::summarise(range = paste0(scales::percent(min(pct)), " to ", scales::percent(max(pct)))) %>% dplyr::pull(range)`.
At the most demanding condition (15°C, high air speed),
`r get_ankle_preference_pct("15C", "high", "Warmer")` preferred warmer
ankles, while at the mildest condition (19°C, low air speed), this
proportion decreased to
`r get_ankle_preference_pct("19C", "low", "Warmer")`.

![Thermal sensation (a) and thermal preference (b) at ankle
level.](figs/thermal_perception_ankle.png){#fig-thermal-perception-ankle
width="75%"}

```{r}
#| label: air-movement-helpers
#| echo: false

# Air movement acceptability summary
air_acceptability_acceptable <- air_movement_acceptability_summary %>%
  dplyr::filter(acceptability == "Acceptable")

air_acceptability_by_sat <- air_acceptability_acceptable %>%
  dplyr::group_by(session_sat) %>%
  dplyr::summarise(
    min = scales::percent(min(pct), accuracy = 1),
    max = scales::percent(max(pct), accuracy = 1),
    .groups = "drop"
  )

# Air movement preference for less
air_pref_less <- air_movement_preference_summary %>%
  dplyr::filter(acceptability == "Less")

air_pref_no_change <- air_movement_preference_summary %>%
  dplyr::filter(acceptability == "No change")
```

### Air movement perception

@fig-air-move-acc-pref shows air movement acceptability and preference
at the ankle level. Despite the localized cooling, air movement at the
ankles was rated acceptable by
`r air_acceptability_acceptable %>% dplyr::summarise(range = paste0(scales::percent(min(pct)), "–", scales::percent(max(pct)))) %>% dplyr::pull(range)`
of participants across all conditions. Acceptability was highest at 19°C
(`r air_acceptability_by_sat %>% dplyr::filter(session_sat == "19C") %>% dplyr::mutate(range = paste0(min, "–", max)) %>% dplyr::pull(range)`)
and lowest at 17°C
(`r air_acceptability_by_sat %>% dplyr::filter(session_sat == "17C") %>% dplyr::mutate(range = paste0(min, "–", max)) %>% dplyr::pull(range)`).
Within each supply air temperature, acceptability tended to decrease
slightly with increasing air speed.

Air movement preference responses indicated that a substantial
proportion of participants would prefer less air movement at the ankles,
particularly at higher air speeds. The proportion preferring less air
movement ranged from
`r air_pref_less %>% dplyr::summarise(range = paste0(scales::percent(min(pct)), " to ", scales::percent(max(pct)))) %>% dplyr::pull(range)`.
At 17°C with high air speed,
`r air_pref_less %>% dplyr::filter(session_sat == "17C", workstation == "high") %>% dplyr::pull(pct_label)`
preferred less air movement, compared to only
`r air_pref_less %>% dplyr::filter(session_sat == "19C", workstation == "low") %>% dplyr::pull(pct_label)`
at 19°C with low air speed. The majority of participants
(`r air_pref_no_change %>% dplyr::summarise(range = paste0(scales::percent(min(pct)), "–", scales::percent(max(pct)))) %>% dplyr::pull(range)`)
preferred no change to the air movement across conditions.

![(a) Air movement acceptability (a) and preference
(b).](figs/air_movement_acc_pref.png){#fig-air-move-acc-pref
width="75%"}

### Skin temperature

```{r}
#| label: tsk-delta-helpers
#| echo: false

tsk_delta_last1_range <- tsk_delta_last1 %>%
  dplyr::group_by(session_sat, workstation) %>%
  dplyr::summarise(mean_delta_last1_c = mean(mean_delta_last1_c, na.rm = TRUE), .groups = "drop") %>%
  dplyr::summarise(
    min_val = round(min(mean_delta_last1_c, na.rm = TRUE), 1),
    max_val = round(max(mean_delta_last1_c, na.rm = TRUE), 1)
  )

tsk_delta_sig_low_med_all <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "medium") %>%
  dplyr::summarise(all_sig = all(p.adj < 0.05, na.rm = TRUE)) %>%
  dplyr::pull(all_sig)

tsk_delta_sig_low_med_max_p <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "medium") %>%
  dplyr::summarise(max_p = max(p.adj, na.rm = TRUE)) %>%
  dplyr::pull(max_p)

tsk_delta_sig_low_high_all <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "high") %>%
  dplyr::summarise(all_sig = all(p.adj < 0.05, na.rm = TRUE)) %>%
  dplyr::pull(all_sig)

tsk_delta_sig_low_high_max_p <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "low", group2 == "high") %>%
  dplyr::summarise(max_p = max(p.adj, na.rm = TRUE)) %>%
  dplyr::pull(max_p)

stopifnot(isTRUE(tsk_delta_sig_low_med_all))
stopifnot(isTRUE(tsk_delta_sig_low_high_all))

tsk_delta_sig_med_high <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "medium", group2 == "high") %>%
  dplyr::mutate(sig = p.adj < 0.05) %>%
  dplyr::filter(sig) %>%
  dplyr::pull(session_sat)

tsk_delta_sig_med_high_p <- tsk_delta_last1_paired_ttest %>%
  dplyr::filter(group1 == "medium", group2 == "high", p.adj < 0.05) %>%
  dplyr::pull(p.adj)

format_p_threshold <- function(p) {
  dplyr::case_when(
    p < 0.001 ~ "p_adj < 0.001",
    p < 0.01 ~ "p_adj < 0.01",
    p < 0.05 ~ "p_adj < 0.05",
    TRUE ~ sprintf("p_adj = %.3f", p)
  )
}
```

Using each participant's skin temperature during the pre-exposure
adaptation period as the baseline, ankle skin temperature during the
final 1 min of exposure was negative under all conditions, ranging from
approximately `r tsk_delta_last1_range$min_val` to
`r tsk_delta_last1_range$max_val`°C. In general, the magnitude of skin
temperature reduction increased with air speed, with medium and high air
speeds typically producing greater cooling than low air speed across all
three supply air temperature conditions. After BH correction, the
differences between low and medium air speed and between low and high
air speed were significant at all three supply air temperatures
(`r format_p_threshold(tsk_delta_sig_low_med_max_p)` and
`r format_p_threshold(tsk_delta_sig_low_high_max_p)`, respectively),
whereas the difference between medium and high air speed was significant
only at `r tsk_delta_sig_med_high`
(`r sprintf("p_adj = %.3f", tsk_delta_sig_med_high_p)`). These findings
suggest that increasing local airflow enhanced the short-term cooling
response of the ankle skin.

![Time course of ankle skin temperature during the formal exposure phase
across the nine experimental conditions: (a) absolute skin temperature;
(b) ankle skin temperture change relative to the adaptation-period
baseline. Shaded bands denote the between-subject mean ± 1
SD.](figs/tsk_p.png){#fig-tsk width="90%"}

## Dissatisfaction with ankle draft

```{r}
#| label: setup-results-data
#| echo: false
#| message: false

# Converting decimals to percentages and rounding for cleaner inline text
res <- model_comparison_d %>%
  mutate(across(c(ppd_liu_mean, dissatisfied_rate, diff), ~ round(.x * 100, 1)))

liu_model_summary <- model_comparison_d %>%
  dplyr::summarise(
    mean_observed = round(mean(dissatisfied_rate) * 100, 0),
    mean_predicted = round(mean(ppd_liu_mean) * 100, 0),
    overprediction_ratio = round(mean(ppd_liu_mean) / mean(dissatisfied_rate), 1)
  )

liu_diff_by_speed <- model_comparison_d %>%
  dplyr::group_by(session_air_flow) %>%
  dplyr::summarise(mean_diff = mean(diff), .groups = "drop") %>%
  dplyr::arrange(dplyr::desc(mean_diff))

liu_max_diff_speed <- liu_diff_by_speed %>% dplyr::slice(1) %>% dplyr::pull(session_air_flow)
liu_min_diff_speed <- liu_diff_by_speed %>% dplyr::slice(n()) %>% dplyr::pull(session_air_flow)
```

@fig-model-comp compares the dissatisfaction rate predicted by the Liu
model with the observations during our experiment. Across all tested
conditions, the existing model consistently overpredicted local thermal
dissatisfaction compared to the observed rates. At an ankle temperature
of 15°C, the discrepancy between predicted and observed rates increased
significantly with air speed: at low flow (0.2 m/s), the difference was
`r res$diff[1]` percentage points (pp), rising to `r res$diff[3]` pp at
high flow (0.6 m/s).

Similar trends were observed at 17°C and 19°C. Notably, for the 19°C
condition at high air speed (0.6 m/s), the predicted PPD was
`r res$ppd_liu_mean[9]`%, while the actual dissatisfied rate was only
`r res$dissatisfied_rate[9]`%, resulting in a mean overestimation of
`r res$diff[9]` pp. On average across all experimental sessions, the
model overpredicted dissatisfaction by `r round(mean(res$diff), 1)` pp.

![Comparison of predicted and observed local thermal dissatisfaction
across varying ankle-level temperatures and air speeds. The dumbbell
plot illustrates the discrepancy (diff) between the Liu model
predictions (ppd_liu_mean) and the actual observed dissatisfied rates
(dissatisfied_rate). Experimental conditions include air speeds of
approximately 0.2 m/s (low), 0.4 m/s (medium), and 0.6 m/s (high).
Points represent mean values, with the segment length indicating the
magnitude of model
overestimation.](figs/model_error.png){#fig-model-comp width="90%"}

Under winter conditions, the observed ankle draft dissatisfaction was
substantially lower than that predicted by the model of
@liuPredictedPercentageDissatisfied2017. Although the model retained
acceptable discrimination on this dataset (AUC =
`r sprintf("%.3f", liu_calibration_stats["C (ROC)"])`), the calibration
results indicated a systematic overestimation under winter conditions,
as shown in @fig-liu-calibration. Across the nine experimental
conditions, the mean observed dissatisfaction rate was
`r liu_model_summary$mean_observed`%, whereas the Liu model predicted an
average of `r liu_model_summary$mean_predicted`%, corresponding to an
overestimation of approximately
`r liu_model_summary$overprediction_ratio` times. The discrepancy was
greatest at `r liu_max_diff_speed` air speed and smaller, but still
evident, at `r liu_min_diff_speed` air speed.

![Calibration curve of the Liu model evaluated on the present dataset
under winter clothing
conditions.](figs/Liumodel_calibrationcurve.png){#fig-liu-calibration
width="50%"}

## Updated model for winter conditions

```{r}
#| label: setup-updated-model
#| echo: false
#| message: false

model_input_coef <- summary(model_input)$coefficients

model_screening <- tibble::tibble(
  term = rownames(model_input_coef),
  estimate = model_input_coef[, "Estimate"],
  std_error = model_input_coef[, "Std. Error"],
  z_value = model_input_coef[, "z value"],
  p_value = model_input_coef[, "Pr(>|z|)"]
)

model_screening_table <- model_screening %>%
  dplyr::filter(term != "(Intercept)") %>%
  dplyr::mutate(
    term_label = dplyr::recode(
      term,
      "v_air_m_s" = "Air speed at ankle",
      "t_air_c" = "Room air temperature",
      "t_supply_c" = "Air temperature at ankle",
      "thermal_sensation" = "Whole-body thermal sensation",
      "turbulence_intensity" = "Turbulence intensity",
      "clothing_typeshort" = "Ankle condition (uncovered vs. covered)",
      "sourcetoby" = "Season"
    ),
    estimate_fmt = sprintf("%.2f", estimate),
    std_error_fmt = sprintf("%.2f", std_error),
    z_value_fmt = sprintf("%.2f", z_value),
    p_value_fmt = dplyr::case_when(
      p_value < 0.001 ~ "<0.001",
      TRUE ~ sprintf("%.3f", p_value)
    )
  ) %>%
  dplyr::select(term_label, estimate_fmt, std_error_fmt, z_value_fmt, p_value_fmt)

model_formula_inline <- model_formula %>%
  dplyr::mutate(
    intercept_fmt = sprintf("%.3f", Intercept),
    v_fmt = sprintf("%.3f", V_coef),
    r2_fmt = sprintf("%.4f", R2),
    formula_text_expanded = sprintf(
      r"(PPD_{AD} = \frac{\exp\left(%s + %sV %s %sTS\right)}{1 + \exp\left(%s + %sV %s %sTS\right)})",
      intercept_fmt,
      v_fmt,
      ifelse(TS_coef < 0, "-", "+"),
      sprintf("%.3f", abs(TS_coef)),
      intercept_fmt,
      v_fmt,
      ifelse(TS_coef < 0, "-", "+"),
      sprintf("%.3f", abs(TS_coef))
    )
  )

model_exposed <- model_formula_inline %>% dplyr::filter(Model == "exposed")
model_covered <- model_formula_inline %>% dplyr::filter(Model == "unexposed")

glmm_exposed_coef <- lme4::fixef(m_glmm_exposed)
glmm_covered_coef <- lme4::fixef(m_glmm_unexposed)
glmm_delta_intercept <- glmm_covered_coef["(Intercept)"] - glmm_exposed_coef["(Intercept)"]

extract_final_model_table <- function(mod, model_label) {
  coef_mat <- summary(mod)$coefficients
  tibble::tibble(
    model = model_label,
    term = rownames(coef_mat),
    estimate = coef_mat[, "Estimate"],
    std_error = coef_mat[, "Std. Error"],
    z_value = coef_mat[, "z value"],
    p_value = coef_mat[, "Pr(>|z|)"]
  )
}

model_final_table <- dplyr::bind_rows(
  extract_final_model_table(m_glmm_exposed, "Ankle uncovered"),
  extract_final_model_table(m_glmm_unexposed, "Ankle covered")
) %>%
  dplyr::mutate(
    term_label = dplyr::recode(
      term,
      "(Intercept)" = "Intercept",
      "v_air_m_s" = "Air speed at ankle",
      "thermal_sensation" = "Whole-body thermal sensation"
    ),
    estimate_fmt = sprintf("%.2f", estimate),
    std_error_fmt = sprintf("%.2f", std_error),
    z_value_fmt = sprintf("%.2f", z_value),
    p_value_fmt = dplyr::case_when(
      p_value < 0.001 ~ "<0.001",
      TRUE ~ sprintf("%.3f", p_value)
    )
  ) %>%
  dplyr::group_by(model) %>%
  dplyr::mutate(
    model_display = dplyr::if_else(dplyr::row_number() == 1, model, "")
  ) %>%
  dplyr::ungroup() %>%
  dplyr::select(model_display, term_label, estimate_fmt, std_error_fmt, z_value_fmt, p_value_fmt)
```

As shown in @tbl-updated-model-screening, in the mixed-effects model
including all candidate predictors, air speed at ankle, whole-body
thermal sensation, and ankle condition were significantly associated
with draft dissatisfaction at the ankle, whereas room air temperature,
air temperature at ankle, and turbulence intensity were not significant
within the range investigated in this study.

```{r}
#| label: tbl-updated-model-screening
#| tbl-cap: "Results of the generalized linear mixed-effects model used for predictor screening."
#| echo: false

knitr::kable(
  model_screening_table,
  col.names = c("Predictor", "Estimate", "SE", "z", "p")
)
```

As shown in @tbl-updated-model-screening, in the mixed-effects model
including all candidate predictors, air speed at ankle, whole-body
thermal sensation, and ankle condition were significantly associated
with draft dissatisfaction at the ankle, whereas room air temperature,
air temperature at ankle, and turbulence intensity were not significant
within the range investigated in this study. Based on these results,
separate mixed-effects models for ankle draft dissatisfaction were
developed for the ankle-uncovered and ankle-covered conditions, using
air speed at ankle and whole-body thermal sensation as predictors.

```{r}
#| label: tbl-updated-model-final
#| tbl-cap: "Generalized linear mixed-effects models for the ankle-uncovered and ankle-covered conditions."
#| echo: false

knitr::kable(
  model_final_table,
  col.names = c("Condition", "Predictor", "Estimate", "SE", "z", "p")
)
```

As shown in @tbl-updated-model-final, the two final models exhibited the
same structure but different coefficients. The intercept of the covered
model was `r sprintf("%.2f", abs(glmm_delta_intercept))` units lower
than that of the uncovered model, indicating a substantially lower
baseline risk of ankle draft dissatisfaction when the ankle was covered
by clothing. In addition, the air speed coefficient was smaller in the
covered model than in the uncovered model, indicating reduced
sensitivity of dissatisfaction to air velocity under ankle-covered
conditions.

For the ankle-uncovered condition, the approximate closed-form equation
derived from the marginal predictions of the mixed-effects model can be
written as:

$$
`r model_exposed$formula_text_expanded`
$$

For the ankle-covered condition, the corresponding approximate
closed-form equation is:

$$
`r model_covered$formula_text_expanded`
$$

Where, $PPD_{AD}$ is the predicted percentage dissatisfied with ankle
draft, $V$ is air speed at ankle (m/s), and $TS$ is whole-body thermal
sensation.

These linear approximations reproduced the marginal predictions of the
mixed-effects models with high accuracy. The corresponding $R^2$ values
were `r model_exposed$r2_fmt` for the uncovered condition and
`r model_covered$r2_fmt` for the covered condition, indicating that the
approximate equations preserved the predictive behavior of the original
mixed-effects models well. The resulting model predictions are shown in
@fig-updated-model.

![Updated ankle draft dissatisfaction model under exposed and covered
conditions. (a) ankle-uncovered conditions, and (b) ankle-covered
conditions.](figs/Model.png){#fig-updated-model width="60%"}

# Discussion

## Comparison with prior research

The core finding of this study is that, under winter clothing conditions
(0.75 clo), the observed dissatisfaction rate caused by ankle-level air
movement is substantially lower than that predicted by the model of
@liuPredictedPercentageDissatisfied2017. Compared with
@schiavonSensationDraftUncovered2016, the overall dissatisfaction rates
in this study are lower. Their study, which involved subjects with
exposed ankles and lower legs wearing approximately 0.6 clo summer
clothing, reported unacceptable rates ranging from 23% to 57%. In
contrast, under ankle-covered conditions (≥0.75 clo) in this study,
airflow unacceptability ranges from 6% to 20%.

These conclusions are consistent with several previous studies.
@zhouDraughtSensationAssessment2022 reported that, in stratified heating
environments, high-top shoes increased ankle airflow acceptability by
approximately 15%, indicating that ankle coverage significantly reduces
sensitivity to airflow. Similarly, Toftum and Nielsen found that, at the
same air velocity, subjects with a cooler overall thermal sensation are
more likely to perceive airflow as uncomfortable, suggesting that when
clothing insulation is higher, the influence of ambient temperature and
its interaction with airflow on perceived discomfort is attenuated
[@toftumDraughtSensitivityInfluenced1996].

Regarding predictor selection, this study explored the inclusion of
additional variables; however, aside from clothing condition, only
overall thermal sensation and air speed (V) were found to have
significant effects on dissatisfaction, consistent with the model
structures reported by @toftumDraughtSensitivityInfluenced1996 and
@liuPredictedPercentageDissatisfied2017.
@chengExperimentalStudyDraftrelated2023 also identified a relationship
between overall thermal sensation and draft rate thresholds in
stratified ventilation experiments, further supporting the use of
overall thermal sensation as a predictor. However, although the dataset
used by @liuPredictedPercentageDissatisfied2017 included both exposed
and covered ankle conditions under summer clothing, no significant
clothing effect was identified. This discrepancy may be attributed to
their use of standard logistic regression, which pooled all observations
across subjects and failed to account for the non-independence
introduced by repeated measures, thereby masking the effect of clothing
as a between-subject variable due to inter-individual variability.

## Mechanisms

```{=html}
<!--
- Why might covered ankles respond differently?
- Thermal insulation of clothing reducing convective heat loss
- Reduced skin exposure and less direct air contact
- Possible psychological factors (feeling "protected")
- Relate to skin temperature findings if included
-->
```

The ankle skin temperature ($T_{\text{sk}}$) results indicate that
winter clothing does not completely block the effect of airflow on the
ankle, but it exerts a substantial attenuating effect. During the final
5-minute steady-state period, a significant reduction in ankle skin
temperature was observed only at 15°C, where medium and high air speeds
were significantly lower than the low speed condition. At 17°C and 19°C,
variations in air speed did not result in significant differences in
skin temperature. In addition, no significant differences were found
between medium and high air speed conditions at any of the three supply
temperatures. That is, only when the environmental contrast is
sufficiently pronounced can the local cooling effect of airflow on the
ankle be detected in steady-state skin temperature. From a
thermophysiological perspective, changes in thermal sensation are more
likely to be associated with rapid variations in skin temperature
[@ChoiInvestigationHumanBody2012; @KenshaloCoolThresholds1968]. In this
study, ankle $\Delta T_{\text{sk}}$ was calculated relative to the
baseline adaptation period. The results show that, across all supply air
temperatures, the differences in $\Delta T_{\text{sk}}$ between low and
medium/high air speed conditions were statistically significant (effect
size $d \approx 0.6\text{--}0.9$). Consistent with this, under the 15°C
condition, local ankle thermal sensation at medium and high speeds was
significantly lower than at low speed, with changes aligned in direction
with the observed skin temperature reductions. Notably, no significant
differences were observed across conditions in terms of air movement
acceptability. In other words, although variations in air speed produced
measurable shifts in both skin temperature and local thermal sensation,
the magnitude of these changes appears insufficient, within the tested
speed range, to generate distinguishable differences in dissatisfaction
rates. From a heat transfer perspective, clothing coverage increases
local thermal resistance and reduces convective heat exchange between
the body and ambient airflow [@lee3DQuantificationMicroclimate2007].
This insulating effect likely suppresses both the magnitude and
variability of skin temperature reduction, making it more difficult for
the human body to cross the thresholds that trigger discomfort.

In addition, ankle coverage may influence airflow acceptability through
psychological pathways, such as altered risk perception or attentional
allocation. Occupants may perceive themselves as “protected” due to
wearing thicker clothing on the lower body, which could increase
tolerance to local airflow. Previous studies have shown that providing
occupants with temperature information or “group perception feedback”
can positively improve thermal comfort evaluations, suggesting that
thermal perception is not solely determined by physical stimuli but is
also modulated by expectations and informational cues
[@zhangGroupPerceptionFeedback2022]. Similarly, research on thermal
expectation indicates that expectation levels themselves influence
thermal perception and evaluation [@schweikerThermalExpectation2020].
Field studies in winter conditions have also identified associations
between perceived control and thermal comfort or satisfaction
[@xuPerceivedControlWinter2021]. These findings suggest that the
subjective perception of being “alreday protected” or “more in control”
due to ankle coverage may enhance tolerance to local airflow. However,
this psychological pathway was not directly measured in the present
study and requires further investigation through controlled manipulation
of coverage conditions alongside measurements of expectation, attention,
and perceived control.

## Implications for building design and standards

```{=html}
<!--
- What does this mean for perimeter heating requirements?
- Under what conditions might perimeter heating be reduced or eliminated?
- Energy and cost implications
- Guidance for practitioners: when to apply winter vs. summer criteria
-->
```

The ankle draft model proposed by
@liuPredictedPercentageDissatisfied2017 has been incorporated into
ASHRAE-55 as a basis for assessing ankle-level draft risk. However, this
model does not distinguish between covered and uncovered ankle
conditions in its formulation. Given that ASHRAE specifies an acceptable
overall thermal sensation range of PMV = −0.5 to +0.5, the current
standard requires ankle level air speed \< 0.22 m/s (corresponding to
\<20% dissatisfied) when TS = −0.5. This constraint may be overly
conservative in practical winter conditions in office buildings.

Based on empirical algorithms considering window surface temperature,
window size, and occupant distance from the façade, the maximum
downdraft velocity at approximately 1 m from a window is around 0.25
m/s. Under milder outdoor conditions or with smaller window areas, peak
velocities are often predicted to fall within the range of 0.1–0.2 m/s
[@lyonsWindowPerformanceHuman1999]. Within this context, for the winter
perimeter conditions examined in this study (air velocity 0.1–0.7 m/s,
supply air temperature 15–19°C), the covered model developed here
suggests that ankle-level dissatisfaction remains within acceptable
limits under many typical winter scenarios. Therefore, the
single-threshold constraint in ASHRAE, based on the Liu model, may lead
to conservative design decisions regarding the sizing or necessity of
perimeter heating or reheat systems. From an energy and system sizing
perspective, perimeter heating systems are often nontrivial in both cost
and energy use. For example, commercial electric perimeter heaters
typically have a rated power density of approximately 150–250 W/ft
(≈490–820 W/m) [@stelproCommercialBaseboardCBF2026]. For a continuous 10
m window perimeter, this corresponds to an installed capacity of roughly
5–8 kW. If “covered” (≥0.75 clo) and “uncovered” (\<0.5 clo) scenarios
are evaluated separately, it may be possible to relax ankle-level air
speed limits under winter clothing conditions, thereby reducing
unnecessary perimeter heating deployment and operation, and unlocking
energy-saving potential and cost reductions.

In practical applications, the model proposed in this study is
consistent with the framework of the ankle draft model in ASHRAE. By
constraining overall thermal sensation within the range of −0.5 to +0.5
and adopting a dissatisfaction threshold of \<20%, acceptable air
velocity limits can be derived.

Within this near-neutral thermal range, the Predicted Mean Vote (PMV)
model provides relatively reliable predictions of overall thermal
sensation. Therefore, PMV can be used as a proxy for overall thermal
sensation during the design or evaluation stage. After specifying the
ankle exposure condition (covered vs. uncovered), the corresponding
model can then be applied to determine the acceptable range of
ankle-level air speed.

## Limitations

```{=html}
<!--
- Laboratory vs. field conditions
- Controlled clothing vs. real occupant clothing variability
- Sample demographics (age range, student population)
- Single body posture (seated)
- Radiant asymmetry from cold windows not simulated
- Air flow direction (from behind only)
-->
```

# Conclusions

# CRediT authorship contribution statement {.unnumbered}

**Tobias Kramer**: Conceptualization, Data curation, Formal analysis,
Methodology, Investigation, Project administration, Software, Writing -
original draft, Writing - review & editing. **Junmeng Lyu**: Formal
analysis, Methodology, Software, Writing - original draft, Writing -
review & editing. **Stefano Schiavon**: Conceptualization, Funding
acquisition, Methodology, Project administration, Supervision, Writing -
review & editing.

# Declaration of competing interest {.unnumbered}

All authors declare XYZ competing interest. The Center for the Built
Environment (CBE) at the University of California, Berkeley, with which
the authors are affiliated, is advised and funded in part by
approximately 50 partners that represent a diversity of organizations
from the building industry, including manufacturers, building owners,
facility managers, contractors, architects, engineers, government
agencies, and utilities.

# Data availability {.unnumbered}

The datasets generated and analyzed during the current study as well as
the full source code of the data analysis workflow are available in the
following GitHub repository: XXX.

# Declaration of Generative AI and AI-assisted technologies in the writing process {.unnumbered}

During the preparation of this work, the authors utilized the chatbot
Claude to enhance the language and readability of the content.
Subsequently, they reviewed and edited the material as required, while
maintaining full accountability for the publication’s content.

# Acknowledgements {.unnumbered}

The Center for the Built Environment at the University of California,
Berkeley supported this research.

# References {.unnumbered}
