---
title: "Thermal comfort responses to ankle-level draft under winter clothing conditions"
authors:
  - name: Tobias Kramer
    affiliations:
      - ref: cbe
    corresponding: true
    email: t.kramer@berkeley.edu
    orcid: 0000-0000-0000-0000
  - name: Junmeng Lyu
    affiliations:
      - ref: cbe
    orcid: 0000-0000-0000-0000
  - name: Stefano Schiavon
    affiliations:
      - ref: cbe
      - ref: arch
      - ref: eng
    orcid: 0000-0003-1285-5682 

affiliations:
  - id: cbe
    name: Center for the Built Environment, University of California, Berkeley, USA
  - id: arch
    name: Department of Architecture, College of Environmental Design, University of California, Berkeley, USA
  - id: eng
    name: Department of Civil and Environmental Engineering, College of Engineering, University of California, Berkeley, USA

filters:
  - authors-block

date: "`r Sys.Date()`"
format:
  docx:
    link-citations: true
    reference-doc: template.docx
    toc: false
    number-sections: true
bibliography: refs.bib
csl: apa.csl
execute:
  echo: false
  warning: false
  error: false
---

```{r load-main-analysis}
#| message: false
#| warning: false
#| paged-print: false
source(here::here("src", "R", "x_analysis.R"))
```

# Abstract {.unnumbered}

<!-- Draft after results are finalized -->

# Introduction

Cold downdraft near windows is a common source of local thermal
discomfort in buildings during winter. When outdoor temperatures drop,
the interior surface of windows—particularly large or poorly insulated
glazing—becomes significantly cooler than the surrounding air. This
temperature differential causes nearby air to cool, increase in density,
and descend, creating a localized airflow pattern that can cause
discomfort at the lower extremities of seated occupants
[@heiselbergDraftComfortReview1994
<!--# Replace reference with BetterBibTex -->]. Building designers
typically address this issue by installing perimeter heating systems,
such as radiators or in-floor heating, to counteract the downdraft
effect and maintain occupant comfort.

The current approach to evaluating ankle draft risk relies on a
predictive model presented in @liuPredictedPercentageDissatisfied2017,
which estimates the predicted percentage dissatisfied (PPD) with ankle
draft as a function of air speed and temperature at ankle level, as well
as whole-body thermal sensation. This model, now incorporated into
ASHRAE Standard 55 [@ashraeANSIASHRAEStandard2023], was derived from
laboratory experiments conducted under cooling conditions typical of
underfloor air distribution (UFAD) and displacement ventilation systems.
Importantly, participants in these foundational studies—including an
earlier investigation by @schiavonSensationDraftUncovered2016—wore
summer clothing with exposed ankles, resulting in clothing insulation
levels of approximately 0.5 clo.

However, building occupants in winter typically wear substantially
different attire. Analysis of the ASHRAE Global Thermal Comfort Database
II [@foldvarylicinaDevelopmentASHRAEGlobal2018] indicates that median
clothing insulation in office environments during heating seasons is
approximately 0.75 clo, with occupants commonly wearing long trousers
and closed-toe shoes that cover the ankle region. This discrepancy
raises an important question: does the existing ankle draft model,
calibrated for summer conditions with exposed skin, accurately predict
discomfort when ankles are covered by clothing?

If the current model overestimates draft risk under winter clothing
conditions, the practical implications are significant. Designers may be
specifying perimeter heating systems that are larger than necessary, or
installing them in situations where they could be avoided entirely. This
has consequences for both first costs and operational energy
consumption, as perimeter heating systems contribute to building energy
use and associated carbon emissions. Conversely, if the model
underestimates risk, occupants may experience discomfort that
compromises satisfaction and productivity.

The objective of this study is to evaluate thermal comfort responses to
ankle-level draft under winter clothing conditions through controlled
laboratory experiments. Specifically, we aim to: (1) characterize local
and whole-body thermal sensation and comfort responses across a range of
supply air temperatures and air speeds representative of window
downdraft conditions; (2) compare observed dissatisfaction rates with
predictions from the existing ankle draft model; and (3) assess whether
the current ASHRAE 55 criteria require adjustment for winter scenarios.
The findings are intended to inform guidance on perimeter heating
requirements and support more energy-efficient building design without
compromising occupant comfort.

# Methods {#sec-methods}

## Experimental design

We conducted a controlled laboratory experiment using a within-subjects
repeated-measures design. Each participant was exposed to nine
experimental conditions in a 3×3 factorial arrangement, varying supply
air temperature (SAT) at three levels (15°C, 17°C, 19°C) and air speed
at ankle level at three levels (low, medium, high; approximately 0.2–0.6
m/s). The SAT levels were selected to represent the range of air
temperatures associated with window downdraft under typical winter
conditions, informed by computational fluid dynamics simulations
conducted as part of a broader research collaboration.

Participants attended four sessions: an introductory session for
consent, training, and baseline measurements, followed by three
experimental sessions of approximately two hours each. Each experimental
session corresponded to one SAT level, with participants experiencing
all three air speed conditions within that session in a counterbalanced
order. Sessions were scheduled on separate days to minimize carryover
effects.

## Climate chamber and apparatus

The experiment was conducted in a climate-controlled chamber at the
Center for the Built Environment, UC Berkeley. The chamber floor area is
approximately 25 m², configured with up to three workstations to allow
simultaneous testing of multiple participants. Ambient room conditions
were maintained at approximately 22°C operative temperature, 50%
relative humidity, and background air velocity below 0.1 m/s, targeting
a whole-body thermal sensation of approximately −0.5 (slightly cool) on
the ASHRAE seven-point scale.

A custom displacement diffuser was used to deliver conditioned air to
the ankle region of seated participants. The diffuser was positioned
approximately 0.7 m behind each workstation, directing airflow toward
the participants' lower legs from behind—simulating the approach
direction of window downdraft. The diffuser was connected to an
independent air handling unit capable of delivering supply air at the
target temperatures and flow rates. Turbulence intensity at the
measurement location was approximately 30%, consistent with values
reported in prior ankle draft studies
[@liuPredictedPercentageDissatisfied2017].

## Environmental monitoring

Air temperature was measured at head-height at each workstation using
calibrated sensors positioned within 0.5 m of participants. Air velocity
at ankle level (0.1 m) was measured using omnidirectional hot-wire
anemometers. Supply air temperature was monitored continuously at the
diffuser outlet. Room-level measurements of relative humidity and CO~2~
concentration were recorded throughout each session.

## Participants

We recruited participants from the university community and local area
through the Experimental Social Science Laboratory (Xlab) platform and
email outreach to individuals who had participated in previous studies
and consented to future contact. Eligible participants were between 21
and 55 years of age, fluent in English, and free from chronic
cardiorespiratory conditions or recent major surgery. Pregnant
individuals were excluded.

<!-- Add: final sample size, demographics summary from analysis -->

## Clothing protocol

To simulate typical winter office attire, participants were instructed
to wear a long-sleeve top, long trousers (ending just above the ankle),
thin socks, and closed-toe shoes. This ensemble targets a clothing
insulation of approximately 0.75 clo, consistent with the median value
observed in office buildings during heating seasons
[@foldvarylicinaDevelopmentASHRAEGlobal2018]. Participants were
permitted to make minor clothing adjustments during sessions (e.g.,
rolling sleeves), and any changes were recorded.

## Experimental protocol

Upon arrival for each experimental session, participants were checked
for elevated temperature and fitted with iButton skin temperature
sensors at standardized locations (lower leg, feet). Participants then
entered the climate chamber and completed a 20-minute adaptation period
at their assigned workstation under baseline conditions.

Following adaptation, participants were exposed to three consecutive
20-minute conditions at different air speed levels, with 20-minute rest
periods between conditions. The sequence of air speed conditions was
counterbalanced across participants and sessions. During each exposure
period, participants completed standardized questionnaires at the end of
each interval.

## Subjective measurements

Thermal comfort was assessed using validated scales consistent with
prior ankle draft research and ASHRAE Standard 55 requirements. Survey
instruments included:

-   **Whole-body thermal sensation**: 7-point ASHRAE scale (−3 cold to
    +3 hot)
-   **Whole-body thermal comfort**: continuous scale
-   **Whole-body thermal preference**: 3-point scale (cooler / no change
    / warmer)
-   **Local thermal sensation at ankles**: 7-point scale
-   **Local thermal preference at ankles**: 3-point scale
-   **Air movement acceptability at ankles**: 5-point scale (very
    unacceptable to very acceptable)
-   **Air movement preference at ankles**: 3-point scale (lower / no
    change / higher)

Surveys were administered electronically via Qualtrics. Participants
were not required to respond to any individual question, though
completion rates were monitored.

## Physiological measurements

Skin temperature was recorded continuously using iButton sensors
(Thermochron, Maxim Integrated) attached with medical-grade
hypoallergenic tape. Sensors were placed on the lateral lower leg
(mid-calf) and feet. Skin temperature data were used to assess local
cooling responses and as physiological correlates of subjective thermal
sensation.

## Statistical analysis and data processing

<!-- Describe analysis approach after methods are finalized -->

# Results

## Participants and environmental conditions

```{r demographics, echo=FALSE}
#| label: tbl-demographics
#| tbl-cap: "Summary of demographic information (mean ± SD) on study participants."

knitr::kable(demographic_d,
             col.names = c('Sex', 'No. Subjects', 'Age [a]', 'Height [m]', 'Weight [kg]', 'Body Mass Index (BMI)')
)

```

```{=html}
<!--
- Final sample size, completion rate, demographics (age, sex distribution)
- Table: Achieved environmental conditions by SAT level and air speed condition
  (mean ± SD for air temp, supply temp, air velocity at ankle, RH)
- Brief statement confirming conditions were achieved as designed
-->
```

```{r env, echo=FALSE}
#| label: tbl-env
#| tbl-cap: "Summary of environmental measurements."

knitr::kable(env_airflow_t)
```

## Thermal sensation and comfort responses

```{=html}
<!--
- Descriptive: Distribution of whole-body and ankle thermal sensation by condition
- Figure: Boxplots or violin plots of thermal sensation (whole-body and ankle) across conditions
- Statistical: Effect of SAT and air speed on ankle thermal sensation (mixed model or repeated measures)
- Key finding: Did ankle sensation decrease (cooler) with lower SAT and higher air speed?
-->
```

![(a) Thermal sensation; and (b) thermal
acceptability.](figs/thermal_perception.png){#fig-thermal-perception}

## Air movement acceptability at ankles

```{=html}
<!--
- Descriptive: Distribution of ankle air movement acceptability by condition
- Figure: Stacked bar chart showing acceptability distribution across conditions
- Statistical: Effect of SAT and air speed on acceptability (ordinal mixed model or Friedman)
- Proportion reporting unacceptable air movement by condition
-->
```

![(a) Air movement acceptability; and (b) air movement
preference.](figs/air_movement_acc_pref.png){#fig-air-move-acc-pref}

## Dissatisfaction with ankle draft

```{=html}
<!--
- Define binary outcome (matching Liu et al. definition)
- Observed dissatisfaction rates by condition (table or figure)
- Comparison with Liu et al. model predictions:
  - Apply Liu et al. model to observed conditions → predicted PPD
  - Compare predicted vs. observed dissatisfaction rates
  - Figure: Predicted vs. observed, or overlay on Liu et al. PPD contour plot
- Key finding: Does current model overestimate/underestimate dissatisfaction under winter clothing?
-->
```

## Updated model for winter conditions

```{=html}
<!--
- Fit logistic mixed model for dissatisfaction (predictors: air speed, supply temp, possibly whole-body sensation)
- Present model coefficients and interpretation
- Figure: PPD contour plot for winter clothing conditions
- Compare model form/coefficients with Liu et al.
- Identify "acceptable" operating zone under winter conditions
-->
```

# Discussion

The primary finding of this study is \[summary of main result regarding
winter clothing effect on ankle draft sensitivity\].

## Comparison with prior research

```{=html}
<!--
- How do results compare to Liu et al. (2017) and Schiavon et al. (2016)?
- Quantify the difference: e.g., "Observed dissatisfaction was X% lower than predicted by the current model"
- Discuss whether this difference is practically meaningful
-->
```

## Mechanisms

```{=html}
<!--
- Why might covered ankles respond differently?
- Thermal insulation of clothing reducing convective heat loss
- Reduced skin exposure → less direct air contact
- Possible psychological factors (feeling "protected")
- Relate to skin temperature findings if included
-->
```

## Implications for building design

```{=html}
<!--
- What does this mean for perimeter heating requirements?
- Under what conditions might perimeter heating be reduced or eliminated?
- Energy and cost implications
- Guidance for practitioners: when to apply winter vs. summer criteria
-->
```

## Relationship to standards

```{=html}
<!--
- How do findings relate to current ASHRAE 55 ankle draft criteria?
- Should the standard include clothing-specific adjustments?
- Recommendations for standards committees
-->
```

## Limitations

```{=html}
<!--
- Laboratory vs. field conditions
- Controlled clothing vs. real occupant clothing variability
- Sample demographics (age range, student population)
- Single body posture (seated)
- Radiant asymmetry from cold windows not simulated
- Air flow direction (from behind only)
-->
```

# Conclusions

## Limitations

# Conclusions

# CRediT authorship contribution statement {.unnumbered}

**Tobias Kramer**: Conceptualization, Data curation, Formal analysis,
Methodology, Investigation, Project administration, Software, Writing -
original draft, Writing - review & editing. **Stefano Schiavon**:
Conceptualization, Funding acquisition, Methodology, Project
administration, Supervision, Writing - review & editing.

# Declaration of competing interest {.unnumbered}

All authors declare XYZ competing interest. The Center for the Built
Environment (CBE) at the University of California, Berkeley, with which
the authors are affiliated, is advised and funded in part by
approximately 50 partners that represent a diversity of organizations
from the building industry, including manufacturers, building owners,
facility managers, contractors, architects, engineers, government
agencies, and utilities.

# Data availability

The datasets generated and analyzed during the current study as well as
the full source code of the data analysis workflow are available in the
following GitHub repository: XXX.

# Declaration of Generative AI and AI-assisted technologies in the writing process

During the preparation of this work, the authors utilized the chatbot
ChatGPT to enhance the language and readability of the content.
Subsequently, they reviewed and edited the material as required, while
maintaining full accountability for the publication’s content.

# Acknowledgements {.unnumbered}

The Center for the Built Environment at the University of California,
Berkeley supported this research. The authors also want to thank
dedicated undergraduate students Heijin Hong, Lana Fiona Lilienstein,
Kevin Kim and Gayle Boyd for their voluntary support throughout the
experimental phase.

# References {.unnumbered}
