```{r}

source(here::here("src", "r", "setup.R"))

```

# Sample Size Calculation

Details:

-   Each participant comes three times (sessions).
-   Each session has a fixed supply-air temperature (SAT): 15°C, 17°C, or 19°C.
-   Within each session, the participant is exposed to three air speed conditions (one at each desk).

So each participant experiences 9 conditions total = 3 SAT levels × 3 air speeds.

In statistical-design terms, that’s a within-subject (repeated measures) 3 × 3 factorial design \* Factor A: SAT with 3 levels (15 / 17 / 19) \* Factor B: Air speed with 3 levels (e.g. low / medium / high)

All conditions are measured on the same people.

## Conceptual model

Acceptability \~ SAT \* Speed + (1 \| Participant)

-   Fixed effects: main effects of SAT and Speed, plus their interaction.
-   Random effects: random intercept for participant (captures repeated measures).

## Alpha and Power

-   Alpha (Type I error): usually 0.05 (two-sided).
-   Power (1 − β): usually 0.80 or 0.90.

Assuming α = 0.05 and Power = 0.80 for now.

## Effect size estimate and sample size calc (based on previous data)

For the 2013 experiment, Q300-1 is the right question. Using `VoteAdj` which seems to be to normalized on a scale from -1 to 1 (previously -4 to 4, vote was divided by 4)

```{r}

# ---------------------------------------------------------
# LOAD DATA
# ---------------------------------------------------------

# Load original dataset from 2013 experiment
data_2013 <- read_csv(here("data", "99-past", "experiment_2013.csv"))


# ---------------------------------------------------------
# STEP 1: Extract THERMAL ACCEPTABILITY responses only
# ---------------------------------------------------------

# Keep only:
# - participant ID (SubID)
# - temperature condition (T1)
# - air speed condition (V1)
# - thermal acceptability responses (VoteAdj -> thermal_acceptability)
# and rename VoteAdj for clarity

acceptability_long <- data_2013 %>%
  mutate(thermal_acceptability = round(VoteAdj, 2)) %>%
  filter(Q1 == "300-1") %>%                   # thermal acceptability question
  select(SubID, T1, V1, thermal_acceptability)

rm(data_2013)

# ---------------------------------------------------------
# STEP 2: Get overall mean acceptability by air speed 
#        (descriptive statistics, not used for power)
# ---------------------------------------------------------

acceptability_means_by_speed <- acceptability_long %>%
  group_by(V1) %>%
  summarize(
    mean_acceptability = round(mean(thermal_acceptability, na.rm = TRUE), 2),
    .groups = "drop"
  )


# ---------------------------------------------------------
# STEP 3: Select ONE temperature (T-Low, T-Mid or T-High) for effect size
# ---------------------------------------------------------

acceptability_T <- acceptability_long %>%
  filter(T1 == "T-High") # assuming that for T-High acceptability differences will be hardest to detect


# ---------------------------------------------------------
# STEP 4: Average within participant × speed
#         (collapse repeated votes per condition)
# ---------------------------------------------------------

acceptability_T_avg <- acceptability_T %>%
  group_by(SubID, V1) %>%
  summarise(
    mean_acceptability = mean(thermal_acceptability, na.rm = TRUE),
    .groups = "drop"
  )


# ---------------------------------------------------------
# STEP 5: Keep LOW vs HIGH air speed only
# ---------------------------------------------------------

acceptability_low_high <- acceptability_T_avg %>%
  filter(V1 %in% c("V-Low", "V-High"))


# ---------------------------------------------------------
# STEP 6: Reshape data so each subject has one row
#         with Low and High in separate columns
# ---------------------------------------------------------

acceptability_wide <- acceptability_low_high %>%
  pivot_wider(
    names_from  = V1,
    values_from = mean_acceptability,
    names_prefix = "speed_"
  )


# ---------------------------------------------------------
# STEP 7: Compute within-person difference scores
#         (High − Low for each participant)
# ---------------------------------------------------------

acceptability_wide <- acceptability_wide %>%
  mutate(
    diff_high_low = `speed_V-High` - `speed_V-Low`
  )


# ---------------------------------------------------------
# STEP 8: Compute effect size for paired comparison
# ---------------------------------------------------------

mean_difference <- mean(acceptability_wide$diff_high_low, na.rm = TRUE)
sd_difference   <- sd(acceptability_wide$diff_high_low, na.rm = TRUE)

dz <- mean_difference / sd_difference     # Cohen's dz

dz_abs <- abs(dz)                          # magnitude only (for power analysis)


# ---------------------------------------------------------
# STEP 9: Power analysis (paired t-test approximation)
# ---------------------------------------------------------

alpha_level <- 0.05
desired_power <- 0.80

power_result <- pwr.t.test(
  d = dz_abs,
  sig.level = alpha_level,
  power = desired_power,
  type = "paired",
  alternative = "two.sided"
)

power_result

```
